\def\filepath{H:/transfer/Math/templates}
%\def\filepath{G:/transfer/Math/templates}

\input{\filepath/packages_book.tex}
\input{\filepath/theorems_with_boxes.tex}
\input{\filepath/macros.tex}
\input{\filepath/formatting.tex}
\input{\filepath/other.tex}

\def\name{Questions}

\input{\filepath/titlepage.tex}

\pagestyle{fancy}
%\addtolength{\headwidth}{\marginparsep} %these change header-rule width
%\addtolength{\headwidth}{\marginparwidth}
\lhead{Questions}
\chead{} 
\rhead{}% 
\lfoot{} 
\cfoot{\thepage} 
\rfoot{} % !! Remember to change the problem set number
\renewcommand{\headrulewidth}{.3pt} 
%\renewcommand{\footrulewidth}{.3pt}
\setlength\voffset{0in}
\setlength\textheight{648pt}

\begin{document}
\maketitle

\chapter{Solved questions}
\section{Analysis}
\begin{enumerate}
\item
What is the relationship between harmonic functions on $\R^2$ and holomorphic functions?
\end{enumerate}

\chapter{Standard questions}
\begin{enumerate}
\item
If $f$ is a Lebesgue integrable function on $[a,b]$, for any $\ep$ there exists a continuous function $g$ such that $\int_a^b |f-g|\,d\mu<\ep$.
\item When is $\fc{d}{dx}\int_a^xf(t)\,dt=f(x)$? When are you allowed to differentiate under an integral sign?
\item If two holomorphic functions agree on a subset with a limit point, then they agree everywhere.
\item Nowhere diff'ble $ab>1+\fc{3\pi}2$ then $W(x)=\sum_{n=1}^{\iy} b^n \cos(a^n x)$.
\end{enumerate}

\chapter{Unsolved questions}
\section{Analysis}
\begin{enumerate}
\item Compare the bounds for harmonic functions with the bounds for holomorphic functions.
\item What definite integrals are amenable to contour integration?
\item Find effective bounds for the Stone-Weierstrass Theorem. See here: \url{http://en.wikipedia.org/wiki/Bernstein_polynomial}.
\item For Fourier expansion: what happens if we choose a different orthonormal basis? Are we similarly doomed?
\item PDE: heat equation. Does it always approach steady-state; in what sense; how does energy decrease?
\item What's a Lebesgue-integrable function that's not equal to a Riemann-integrable function a.e.? (Try Cantor set like thing)
\end{enumerate}


\section{Number theory}
\begin{enumerate}
\item
Try to prove the following statement for as small a $y$ as possible: $\pi(x+y)-\pi(x)\sim \fc{y}{\ln x}$.
\item
Find the asymptotics of the numbers $n<x$ that can be expressed as the sum of 2 squares.
\item 
\end{enumerate}

\section{Geometry}
\begin{enumerate}
\item
\end{enumerate}

\chapter{Research questions}
\section{Category theory}
\begin{enumerate}
\item
In many cases, we say there's one obvious natural definition of a map. 
How can you get a computer to find the ``natural definition of a map"? I.e., what do we mean by obvious natural definition in computer science terms?
\end{enumerate}

\section{Number theory}
\begin{enumerate}
\item
How small of a primitive root can you find? Given a multiplicative subgroup of $\F_p$, what is its density in $\F_p$ up to $x$? (Something to do with character sums?)
\item Effective bounds for equidistribution.
\item Effective Skolem-Mahler-Lech \url{http://terrytao.wordpress.com/2007/05/25/open-question-effective-skolem-mahler-lech-theorem/}
\end{enumerate}
\section{Geometry}
\begin{enumerate}
\item
Consider a ball rolling in a polygonal pool table. Suppose its trajectory is not periodic. Is it ergodic? What about "unilluminable points" in 3 dimensions? (Try the ergodicity for a square room first.)
\item
Ravi's problem.
\end{enumerate}

%\input{chapters/1.tex}
 
%\bibliographystyle{plain}
%\bibliography{refs}
\end{document}