\def\filepath{C:/Users/Owner/Dropbox/Math/templates}
%\def\filepath{C:/Users/holden-lee/Dropbox/Math/templates}
%\def\filepath{G:/transfer/Math/templates}

\usepackage{showkeys}
\input{\filepath/packages_book.tex}
\input{\filepath/theorems_with_boxes.tex}
\input{\filepath/macros.tex}
\input{\filepath/formatting.tex}
\input{\filepath/other.tex}
\input{\filepath/theorem_num.tex}

\def\name{Math threads}

\input{\filepath/titlepage.tex}

\pagestyle{fancy}
%\addtolength{\headwidth}{\marginparsep} %these change header-rule width
%\addtolength{\headwidth}{\marginparwidth}
\lhead{Math threads}
\chead{} 
\rhead{}% 
\lfoot{} 
\cfoot{\thepage} 
\rfoot{} % !! Remember to change the problem set number
\renewcommand{\headrulewidth}{.3pt} 
%\renewcommand{\footrulewidth}{.3pt}
\setlength\voffset{0in}
\setlength\textheight{648pt}

\begin{document}
\maketitle

\tableofcontents

\chapter*{Introduction}

The goal of this document is to
\begin{enumerate}
\item
Keep track of my own thoughts about math, and in particular save partial progress.
\item 
Provide links to useful references.
\item
Write up open-source notes about subjects that I learn.
\end{enumerate}

\section{Math thoughtstream}

I'm going to catalogue all the math-related thoughts I have every day here (like a math thoughtstream \url{http://www.workflowy.com}). (This is inspired partly by Jacob Cole's idea of a ``curiosity thread," \url{http://curiositythread.tk/}.) I find myself running into a few problems when I learn math:
\begin{enumerate}
\item
I note interesting questions, and then forget about them if they don't seem pressing enough, or I misplace them.
\item
I take down notes, and don't bring the notes when I need them (or spend too long looking for them). 
\item There's lots of math topics, articles, blogs, etc. I want to read, but because of limited time, have to postpone, and I need a centralized place to keep the links so I can find them later.
\item
Sometimes I read things, hold them in short-term memory, think I understand them, and then several months later I have to reread it to understand it. This is suboptimal---I want to force myself to write a summary, no matter how bare-bones, to save ``partial progress in understanding," so if I work on something, going back to it I can pick up where I left off in understanding.
\end{enumerate}
I've been ``too busy" learning math to keep a blog, so this will be a substitute---I intend to write up the more interesting/important stuff in nicer form.

%More on (4):
%One thing that frustrates me with learning math is forgetting things (usually from not learning them properly in the first place). I'm currently keeping a document (a "thread") where I summarize what I learn every day, and problems I solve including motivations/why things that didn't work didn't work. I'm writing out index cards for the concepts I learn, so I can see all the important information at a glance. I'm starting to put in summaries and motivations in my LiveTeXed notes (see the ones that have my name by them here: \url{https://docs.google.com/document/d/1KZqXRf1xXlqQh_G-yG1DwgpUEmCo0DLVxe9_33eInQc/edit?usp=drive_web}) to make them more helpful.
%All this takes a lot more time, but I want to make my knowedge more ``additive": by recording my thoughts (and partial progress) in figuring out something, I can look at my notes again and return to that level of understanding more quickly.

\section{What's different}
\begin{enumerate}
\item
This is not a blog; it is a thoughtstream. (1) In a blog, I could write up math articles nicely, but it's harder to show/keep track of where they fit in my personal ``big picture" of the math that I know. See my blog at \url{http://holdenlee.wordpress.com}. (2) I don't filter out material, other than using labels (see Organization, below), as I don't think there's harm in making this available.
\item
It is not intended to stand on its own: I make copious references to books, articles, and websites. Part of it is notes made while reading books. (The idea is like in~\url{http://mathbooknotes.wikia.com/wiki/Math_Book_Notes_Wiki}).
%\item
%It is intended to be in one place. I'm personally very bad at browsing a lot of different places (though I am warming up to places like math stackexchange and stackoverflow); it helps me a lot for my thoughts to be in one place.
\end{enumerate}
Why LaTeX/pdf? For me it is the fastest way to write math on the computer. Other formats don't allow LaTeX macros.

\section{Organization}

To help a reader navigate, I'll use the following labels. All numbers are rough guidelines only.
\begin{enumerate}
\item
Readability: on the scale of 1-5, how easy it might be for someone else to read. 1 means not at all (probably lots of random notes I wrote up trying to understand something), 3 means somewhat comprehensible (but probably not finished, missing chunks, etc.), and 5 means ``this could go in a book."
\item
Precision: on the scale of 1-5, how much I explicitly go through the theorems and proofs. 1 means that I only try to write things in a big-picture kind of way (that may be misleading/inaccurate) so take these with a grain of salt.
\item 
Recommendations: [R] for books/references I highly recommend.
\end{enumerate}
At the start of each section I'll (eventually) have a guide to that section, including: what I've learned and what I'm learning (in red), what I'd like to learn, and references/links.

\section{Contact}

I would love it if more people joined in! For a quick start, use the source code and delete the text (contact me if you need help); if enough people are interested we can start a group.

Email me at~\url{mailto:holdenlee@alum.mit.edu} or~\url{mailto:hl422@cam.ac.uk} to talk about math, or to talk about meta things like how to improve the organization of thoughtstreams, collaboration in math, etc.
 
%My eventual vision is a lot of open-source, remixable math notes.

\chapter{Threads organized by date}

\section{January 2014}
\begin{enumerate}
\setcounter{enumi}{26}
\item %27
(Analysis) Learned Fourier analysis
(Computation) Read Ch. 9 of Quantum computing since Democritus (Aaronson).
\item %28
(Computation) Learned how to use Isabelle.
(Algebra) Chapter 4 of Schemes (Vakil).
(Computation) Continued Aaronson.
\hyperlink{comp/qc/1-28-14}{Notes.} Did QComp problems 1 (Bernstein-Vazirani) and 3 (Simon).
(Number theory) Went over Van der Corput lemma and corollary (equidistribution of $f(j)$ given by $\De_r f(j)$, \hyperlink{nt/ac/weyl-equid}{notes}).
(Logic) Thought about G\"odel Incompleteness. (Have lots of questions on this.)
\item Went over combo, percolation, distributions material. Read Ch. 11 of Aaronson. (Logic) Did exer. 27.
\item (Probability) Reading Ch. 4 of Probability on graphs. 
Wrote blog post on mathematical compression.
\item (Probability) Finish most of Ch. 4.
\end{enumerate}

\section{February 2014}
\begin{enumerate}
\item
(Analysis) Did distributions problems 1-7, parts of 11, 13.
(Probability) Did problem 9 of percolation.
(Algebra/geometry) Read Ch. 4 of Vakil and worked on exercises.
\item
(Algebra/geometry) Worked through Ch. 5 (skimming parts) (Logic) Read part of ``Reflection in probabilistic logic."
\item
(Computation) Essentially finish Quantum computation examples 1. (Algebraic geometry) Start writing summary. \hyperlink{geo/ag/summary}{Notes}.
\item
(Combinatorics) Worked on Adam's graph theory problem (without success). (Algebra/geometry) Start Ch. 8.
\item
(Algebra) Polynomial method: cell decomposition, Szemeredi-Trotter, Bezout.
\item
(Combinatorics) Solved Adam's graph theory problem.
\item -
\item -
\item (Combinatorics) Reading O'Donnell's thesis on noise sensitivity
\item (Analysis) Worked on distributions examples, typed (in distributions doc), read 3.1 of Stein-Shakarchi.
\item (Logic) Met with Prof. Forster to ask questions. (Percolation) Started examples sheet. (Computation) Worked on exercise for IFV.
\item (Combinatorics) Wrote up Bonami-Beckner. Read more on noise sensitivity. (Algebra) Started AMiIT problems.
\item (Analysis) Distributions: did most of examples. See distributions notes.~\url{https://dl.dropboxusercontent.com/u/27883775/math\%20notes/part_iii_distributions.pdf}
Read parts of Stein-Shakarchi Ch. 2 and 3 (Hilbert transforms, inequalities using convexity).
\item (Combinatorics) Wrote up Fourier spectrum small $\implies$ $II\to 0$ argument, and iterated majority. (Computation) Wrote up QComp \hyperlink{comp/qc/ex1}{exercises} and index-carded. (Question \url{http://tex.stackexchange.com/questions/160500/using-qasm2circ})
\item (Combinatorics) Did most of Extremal combo sheet. See part iii combo~\url{https://dl.dropboxusercontent.com/u/27883775/math\%20notes/part_iii_combo.pdf}
\item (Probability) Percolation: worked on \hyperlink{prob/perc/ex1}{examples 1}.
\item (Computation) Worked on IFV exercise.
\item ?
\item (Computation) Index-carded 18.404, read quantum computation
\item (Computation) Continued quantum computation
\item (Combinatorics) Worked on noise sensitivity.
\item -
\item TMS lecture---maths of climate, dependent type theory. (Computation) Index-carded quantum computation.
\item (Combinatorics) Did examples 1. See~\url{https://dl.dropboxusercontent.com/u/27883775/math\%20notes/part_iii_combo.pdf}
\item (Combinatorics) Skimmed some noise sensitivity (VI). 
\item (Algebra) Polynomial method. (Analysis) Distributions
\item ?
\item (Computation) Caught up notes
\end{enumerate}
\section{March 2014}
\begin{enumerate}
\item (Combinatorics) Noise sensitivity reading.
\item -
\item (Combinatorics, Computation) Noise sensitivity, boolean functions
\item (Combinatorics) Did Problem 1 in ES1. NS reading. (Algebra) Polymethod. 
\item (Probability) Reviewed and index-carded percolation.
\item ?
\item (Combinatorics) Started examples sheet. Index-carded LLL and correlation inequalities.
\item (Analysis) Started ch. 3 of FAoNF, did some distribution examples. (Algebra) Polynomial method---read Ch. 2 of Dvir's notes. Read Terry Tao's blog entries on PM.
\item (Computation) Started Ch. 11 of Nature of Computation. (Combinatorics) Noise sensitivity, did exercises to understand simple examples.
\item (Combinatorics) Noise sensitivity. (Computation) Reading from NoC.
\item
(Number theory) Analytic class number formula
\item
Part III talks. (Number theory) Continue.
\item Part III talks. (Number theory) 
\hyperlink{nt/ant/aant}{analytic class number formula writeup}, $x^2+x+p$ problem
\item (Number theory) Wrote up ACNF.
\end{enumerate}

\section{April 2014}

\section{May 2014}
\begin{enumerate}
\item
\item
\item
\item
\item
\item
\item
\item
\item
\item
\item
\item %12
Revise for Ramsey.
\item Revise for Elliptic Curves
\item Elliptic
\item Elliptic (formal groups, local fields, torsion group, Kummer thery, Galois cohomology)
\item Elliptic (Mordell-Weil with computation, heights)
\end{enumerate}

%\input{thread/thread_11-13.tex}

%%%%%%%%%%%%%%%%

\input{Analysis/analysis.tex}


\input{Combinatorics/combo.tex}

\input{Computation/computation.tex}
%\section{Computation}

\section{Unsorted}

\fixme{[Read: 1, Prec: 1]}

Q: for what kind of existence questions is it true that either ``$A$ exists" or ``$A$ doesn't exist" consistently in all axiom systems (that include ZF)? 

I.e., ``P$=$NP" under one axiom system and not under another is impossible, right? Try to enumerate all programs, and try to write proofs that a program solves 3-SAT... what if a program solves 3-SAT but it is unprovable that it does? Well, still it either has to solve it or not solve it, and the existence of one under any axiom system gives a program that works under any axiom system, right? (Do I have my logic right?) If P$=$NP is there necessarily a certificate?

On the other hand, it is possible that P$\ne$NP and there is no way to prove it, i.e., it's undecidable. 

\input{Geometry/geometry.tex}

\input{Probability/probability.tex}
\input{Statistics/stat.tex}


\input{Logic/logic.tex}

\input{"Number\space Theory/number.tex"}

\input{thread/index_cards.tex}

\chapter{Questions}

%\section{Review questions}
%
%\subsection{Analysis}
%\begin{enumerate}
%\item
%Define regularity. Why is it important?
%\item
%Lebesgue measure is defined on the Borel algebra on $\R^n$. Can it be extended beyone the Borel algebra? If so, how far?
%\end{enumerate}

\section{Interesting questions}

Combinatorics
\begin{enumerate}
\item
18.821 Project II: sequences avoiding solutions to certain equations. Hypothesize: it has a nice structure iff it's partition regular? @imre
\item 
What happens when we avoid higher-degree equations? @imre
\end{enumerate}

\section{Research questions}

\section{Things I'm interested in}
Too vague to make interesting questions yet at this point.

Big things: ((FF) means far-fetched.)
\begin{enumerate}
\item
What Strogatz has done with small-world networks. How can we characterize in an abstract and simple way the way that the following are organized (and the way they evolve)?
\begin{enumerate}
\item neurons in the brain
\item the web of knowledge, specifically, mathematical theorems (cf. Jacob's math friend)
\end{enumerate}
Related stuff
\begin{enumerate}
\item Graph theoretic ideas of order and disorder, ex. optimal bounds for Szemeredi's Theorem. Are there more flexible versions of these kinds of theorems for certain families of graphs, for example graphs of knowledge above?
\item defining metrics
\end{enumerate}
\item 
Mathematical formulations of complexity, intelligence and information (Kolmogorov complexity, AIXI, ``consciousness," machine learning, etc.). In particular, defining complexity (see scholarpedia article).
\begin{enumerate}
\item Theoretical computer science: computational complexity, P vs. NP, etc. How far does ``philosophy" go into this?
\item Can we define complexity for something like cellular automata?
\item Mathematical theorem proving. Representing mathematical knowledge in useful form. (ex. analogies, parallels between different arguments.
\item Information compression. The Hutter prize.
\item Defining layers. Programs on several levels. Are genes the same way?
\item (FF)
Is there a way to characterize what makes a good story mathematically? How to fit some part into another, as in Genesis? Better math than PCA for ConceptNet?
(Tensor products??)
\end{enumerate}
\item Open problems that are cool and simply stated (and that probably don't require the kind of machinery that hard number theory problems typically require), Erd\H os-like problems.
\begin{enumerate}
\item
\url{http://www.openproblemgarden.org/op/limiting_subsequence_sums_in_z_n_x_z_n}, 
\url{http://www.openproblemgarden.org/op/quartic_rationally_derived_polynomials} 
\url{http://www.openproblemgarden.org/op/a_discrete_iteration_related_to_pierce_expansions} 
\url{http://www.openproblemgarden.org/op/snevilys_conjecture} 
\url{http://en.wikipedia.org/wiki/Erd\%C5\%91s_conjecture}
\url{http://en.wikipedia.org/wiki/Erd\%C5\%91s\%E2\%80\%93Straus_conjecture\#CITEREFElsholtzTao2011}
\url{http://www.openproblemgarden.org/op/erdos_faber_lovasz_conjecture}
\url{http://www.math.ucsd.edu/~erdosproblems/All.html}
\end{enumerate}•
\item Randomness in arithmetic.
\begin{enumerate}
\item
Arithmetic dynamics.
\item
Questions of equidistribution in number theory.
\item Additive combinatorics. Randomness and structure.
\end{enumerate}
\item The kind of stuff Diaconis does: a way of doing applied math by asking lots of questions, talking to people, browsing. Ex. dynamics of card-shuffling.
\end{enumerate}

Fuzzy

\url{http://statweb.stanford.edu/~cgates/PERSI/papers/scho1.pdf} Convolution $\lra$ group expansion?
\url{http://statweb.stanford.edu/~cgates/PERSI/Courses/Phil166-266/index.html}
\chapter{To read}

\begin{enumerate}
\item
Stuff on complexity
\begin{enumerate}
\item 
\url{http://www.scholarpedia.org/article/Algorithmic_information_theory}
\item
\url{http://www.scholarpedia.org/article/Neuroanatomy}
\end{enumerate}
\item P vs. NP, Borel determinacy
\item Randomness in number theory
\begin{enumerate}
\item
 equidistribution, thin groups, Ramanujan graphs in number theory, quantum arithmetic chaos. Peter Sarnak's stuff.
\end{enumerate}
\item Elementary number theory:
\begin{enumerate}
\item \url{http://en.wikipedia.org/wiki/Pythagorean_triple\#Parent.2Fchild_relationships}
\item Continued fractions
\end{enumerate}
\item Number theory
\begin{enumerate}
\item Tate's Thesis
\item
http://mathoverflow.net/questions/52576/zeta-function-zero-density-theorems
\item Transcendental number theory
\end{enumerate}•
\item Cool!
\begin{enumerate}
\item
Persi Diaconis's paper
\end{enumerate}
\item Connections
\begin{enumerate}
\item Stats: on the computability properties of finding the right sparse vector
\item Functional analysis and quantum mechanics
http://www.physicsforums.com/showthread.php?t=524351
\end{enumerate}
\item Books
\begin{enumerate}
\item
An epsilon of room.
\item Kolmogorov complexity
\item Nonlinear dynamics and chaos, Strogatz.
\end{enumerate}
\end{enumerate}

General:
\begin{itemize}
\item
\url{http://terrytao.wordpress.com/}
\item
\url{http://blog.scholarpedia.org/2013/11/04/wikifying-scholarly-canons/}
\item
\url{http://jhnpappas.wordpress.com/}
\end{itemize}


\printindex

%\input{chapters/1.tex}
 
%\bibliographystyle{plain}
%\bibliography{refs}
\end{document}