Reference: 
\begin{enumerate}
\item
Chapter 4.4 and 4.9 in \cite{Clarke2009}.
\item
Barron's original paper and followup, \cite{Barron1993}, \cite{Barron1994}.
\end{enumerate}•

\section{The curse of dimensionality}
%The intermediate tranche is where the finite-dimensional methods confront
%the Curse of Dimensionality on their way to achieving good approximations to the
%nonparametric setting.
In statistics and machine learning, the standard setting is the following. Given $(x_i\in B,y_i)$, a function $f_\te(x)$, find parameters $\wt{\te}$ such that $f_{\wt\te}(x_i)\approx y_i$. What is our metric of success? We want an algorithm that minimizes the \vocab{mean integrated square error} (let $\wt f=f_{\wt \te}$)
\[
\text{MISE}%(\wt f) 
= \E_{(x_i,y_i)} [\ve{\wt f-f}_2^2] 
\]
where $\ve{f}_2=\int_B f(x)^2\,d\mu(x)$. The expected value is over independent random $(x_i,y_i)$, and the integral is with respect to the probability distribution on the samples $x_i$. (We assume there is a distribution on the $x$'s.) %(or at least, random $y_i$ if we don't assume a distribution of the $x_i$).

The \vocab{curse of dimensionality} is a general phenomenon where estimates degrade with the number of dimensions. Consider a class of models $f_{p,\te}:\R^p\to \R$. For useful classes of models, the MISE typically increases superlinearly in $p$, and the number of data points required also increases rapidly.

Now let's look at the setting of neural nets.

\begin{df}
A sigmoidal function is a differentiable function $f$ on $\R$ with $f'>0$, $\lim_{x\to -\iy}f(x)=0$, and $\lim_{x\to \iy}f(x)=1$.
\end{df}

We have the following.
\begin{pr}
Let $\phi$ be sigmoidal. 
Every continuous function on a bounded set $B\subeq \R^p$ can be approximated to arbitrary precision by a linear combination of $\phi(a\cdot x+b)$. 
\end{pr}
Such a combination is represented by a (1-layer) neural net where 
\begin{itemize}
\item
the input layer has $p$ nodes, i.e., represents an element of $\R^p$,
\item
the hidden layer has some number of nodes,
\item
the output node is a linear combination of hidden layer nodes.
\end{itemize}
(We're trying to approximate a function rather than make a decision, so we don't take a threshold function at the output.)

A natural question is how well can such a neural net approximate an arbitrary continuous function? We'll give a precise answer, depending on the regularity of $f$ and the size of the hidden layer we allow, but {\it not the dimension $p$}.
Barron's theorem tells us that ``neural nets evade the curse of dimensionality" in the following sense.

{\it The best 1-layer neural net approximations do not get worse as $p$ increases.}

Since the number of hidden nodes required to get $\ep$ approximation does not depend on $p$, and the number of parameters is $O(\pat{hidden nodes}p)$, the number of parameters we need is linear in $p$ rather than superlinear.

Note we are not saying anything about an {\it algorithm} to find the best approximation. The loss function is in general not convex so it's unclear whether gradient descent will actually find the approximation that Barron's Theorem gives. If we care about the actual $\wt f$ that we estimate from sample data $(x_i,y_i)$, then we have to include the estimation error:
\[
\text{MISE} \le \ve{f-f_{\text{best}}}_2+\E_{(x_i,y_i)} \ve{f_{\text{best}}-\wt f}_2.
\]
We'll discuss the first term in the Section~\ref{sec:barron}, and the second term in Section~\ref{sec:barron2}.


%The number of hidden nodes required to get an approximation of  $O\prc{r}$

\section{Barron's Theorem}
\label{sec:barron}
The Fourier transform of $f:\R^p\to \R$ is
\[
\wh f(\om)=\rc{(2\pi)^p}\int_{\R^p} f(x)e^{-i\om\cdot x}\,dx.
\]
The Fourier inversion formula is
\[
f(x) = \int \wh f(x)e^{i\om\cdot x}\,dx.
\]
When $B$ is the unit ball, our measure of smoothness will be the following:
\[
\ve{\wh{f'}}_1 = \ve{\om \wh f}_1 = \int_{\R^p} |\om \wh f(\om)|\,d\om.
\]
More generally, for an arbitrary bounded set $B\subeq \R^p$ %conditions,
let $|\om|_B=\sup_{x\in B} |\om \cdot x|$. When $B=B_0(1)$ is the unit ball, this is simply $\ve{\om}_2$. %We'll consider the 
In the general setting our smoothness measure is 
\[
\ve{f}_B^* :=\int_{\R^p} |\om|_B |\wh f(\om)|\,d\om.
\]

Let $\Ga_B$ be the set of functions on $B$ where the Fourier inversion formula holds after subtracting out the mean,\footnote{for example, it includes all smooth ($C^{\iy}$) functions and more generally, all $L^1$ functions on $B$ whose Fourier transform is also $L^1$}
 %conditions?
\[
\Ga_B=\set{f:B\to \R}{\forall x\in B, f(x)=f(0) + \int (e^{i\om \cdot x}-1)\wh f(\om)\,d\om}
\]
Let $\Ga_{B,C}$ be the subset with smoothness $\le C$:
\[
\Ga_{B,C}=\Ga_B \cap \{\ve{f}_B^*\le C\}.
\]
The quality of the approximation will depend on how large the phases of $f$ are. We'll see in the proof where the norm $\ve{f}_B^*$ arises.

\begin{thm}[Barron]
Let $B\subeq \R^p$ be a bounded set, $\mu$ a probability measure on $B$, and $\ep>0$.
Let $f\in \Ga_{B,C}$ and $\phi$ be sigmoidal. There exists
\[
f_r=\sum_{i=1}^r c_i\phi(a_i\cdot x+b_i)
\]
with $\sum_{i=1}^r |c_i|\le 2C$ such that 
\[
\ve{f-f_r}^2 =\int_B (f(x)-f_r(x))^2\,\mu(dx) \le \fc{(2C)^2}{r}+\ep.
\]
\end{thm}
%The theorem also holds with an arbitrary measure $\mu$.
We'll just consider the case when $\mu$ is uniform on $B$, but in general, the proof goes through the same way (with a bit more care). %For simplicity we just consider normal Lebesgue measure.

This means that the number of parameters required to get an approximation of $\ep$ is $(p+2)r = (p+2)\fc{(2C)^2}{\ep}$, which is linear in $p$ rather than superlinear.

The idea of the proof is the following. 
\begin{enumerate}
\item
Show that $f$ is in the closed convex hull of the $\phi$'s. We break this into several inclusions which we show one at a time:
\bal
\{\ve{f}^*\le C\} &\stackrel{(3)}{\subeq}
\oconv \ub{\set{\fc{\ga}{|\om|_B}(\cos (\om \cdot x+b)-\cos b)}{\om \ne 0, |\ga|\le C}}{=:G_{\cos}}\\
& \stackrel{(2)}{\subeq}
\oconv \ub{\set{cH(a\cdot x +b)}{c\le 2C, |a|_B=1, |b|_B\le 1}}{=: G_{\text{step}}}\\
& \stackrel{(1)}{\subeq}
\oconv \ub{\set{c\phi(a\cdot x+b)}{c\le 2C}}{G_{\phi}}
\end{align*}
where $H$ is the step function $1_{x\ge 0}$. 
We explain the inclusions.
First, the exact form of $\phi$ doesn't matter: all we need about $\phi$ is that it can approximate step functions arbitrarily well. ($\phi$ sigmoidal gives us this.) 
\item Second, we write the step functions in terms of a standard basis, namely the Fourier basis. 
\item Third, we write out the Fourier expansion of an arbitrary regular $f$ to show that $f$ is in $G_{\cos}$.
\item
Next, we use a general fact: If $A$ is convex and $f\in \oconv A$, then $f$ is close to a small combination of elements of $A$. This in fact holds in any Hilbert space. The proof is by writing $f$ as a linear combination, and then sampling the functions with probabilities given by the coefficients.

Thus $f$ being in the convex hull of the $\phi$'s gives us that $f$ is close to a small combination of them.
\end{enumerate}

\begin{proof}
\begin{enumerate}
\item
Without loss of generality, $\phi$ is centered at 0. Then 
\[
\phi(k(a\cdot x+b))\to H(a\cdot x+b)
\]
for $x\ne 0$ so $G_{\text{step}}\subeq \ol{G_{\phi}}$.
\item
We relate $H$ to the the Fourier basis: $G_{\cos}\subeq \oconv (G_{\text{step}}^\mu)$. 
We can do this easily because each $\cos(\om\cdot x+b)-\cos b$ is 1-dimensional. (This is why Fourier transforms are useful in this proof: $\om\cdot x+b$ is a projection of $x$ onto the $\om$ direction.) 

Let $g(y)=\cos(|\om|_By+b)-\cos(y)$. Let $x_{-k},\ldots, x_k$ be a partition of $[-1,1]$ such that $g$ changes by $<\ep$ on each interval, we can approximate $g$ to within $\ep$ at every point by the sum
\[
\sum_{i\ge 0} (g(x_{i+1})-g(x_i))1_{\ge x_i} +\sum_{i\le 0} (g(x_{i-1})-g(x_i)) 1_{\le x_i}.
\]
The sum of coefficients is 
\[\sum_i |g(x_{i+1})-g(x_i)|\le \int |g'|\,dx\le 2|\om|_B.\]
Now substitute $y=\fc{\om}{|\om|_B}\cdot x$ to get the approximation of $\cos(\om\cdot x+b)$ by a linear combination  with sum of coefficients $2|\om|_B$, i.e., an approximation of 
$\fc{\ga}{|\om|_B}(\cos(\om\cdot x + b) - \cos b),\om \ne 0,|\ga|\le C$ by a linear combination of $H$'s with sum of coefficients $2C$. \footnote{For an arbitrary measure, there is an extra step where we show that we can restrict $-b$ to the continuity points of the measure $\mu$.}
\item
When is $f\in \oconv (G_{\cos})$? We show $\{\ve{f}^*\le C\}\subeq \oconv (G_{\cos})$.
Use Fourier inversion. Write the Fourier transform in polar form as $\wh f=|\wh f|e^{i\te(\om)}$: 
\bal
f(x)-f(0)&= \int \wh f(\om) (e^{i\om \cdot x}-1)\,d\om \\
&= \int |\wh f| e^{i\te(\om)} (e^{i\om \cdot x}-1)\,d\om\\
&= \int |\wh f|(\cos(\om \cdot x+\te(\om))-\cos( \te(\om)))\,d\om&\text{taking real part}\\
&= \int |\wh f||\om|_B \rc{|\om|_B} (\cos(\om \cdot x+\te(\om))-\cos( \te(\om)))\,d\om.
\end{align*}
Hence, so long as $\int |\wh f||\om|_B\le C$, $f$ is in a combination of functions in $G_{\cos}$ with sum (integral) of coefficients $\le C$. (The integral is in the closure of the convex hull because it can be approximated as a Riemann sum.)
\item
We show the following.
\begin{lem}
Let $G$ be a bounded set in a Hilbert space, where every element has norm $\le b$. (For example, $G\subeq L^2(B)$.) Let $f\in \oconv(G)$. Then for every $r$,
\[
\inf_{f_r=\sum_{i=1}^r c_ig_i,g_i\in G, \sum c_i=1} \ve{f-f_r}^2 \le \fc{b^2-\ve{f}^2}{r} \le \fc{b^2}{r}.
\]
(The infimum is taken over all convex combinations involving $r$ functions.)
\end{lem}
\begin{proof}
Since $f\in \oconv(G)$, for all $\ep$, there exists $f^*$ in the following form that is $\ep$ away from $f$:
\[
f\approx_\ep f^*=\sum_{i=1}^m c_i g_i^*.
\] 
Let $g$ be a random variable such that 
\[
g=g_i^*\text{ with probability } \fc{c_i}{\sum_{j=1}^m |c_j|}.
\]
Let $g_1,\ldots, g_r$ be $r$ independent draws, and let $f_n$ be the average,
\[
f_r = \rc r\sum_{i=1}^r g_i.
\]
Then (since $f_r$ is the average of $r$ variables distributed as $G$ and $f^*=\E g$)
\bal
\E\ve{f_r - f^*}^2 &= \rc r\E\ve{g-\E g}^2 \\
&=\rc r [\E(g^2)-(\E g)^2] \\
&\le \rc r(b^2-\ve{f}^2).
\end{align*}
\end{proof}
Finally, apply the lemma to 
\[
f\in \oconv\set{c\phi(a\cdot x+b)}{c\le 2C},
\]
noting that the norms of the $\phi$'s are $\le 1$ since $\mu$ is a probability measure.
\end{enumerate}
\end{proof}

\section{Barron's Theorem: Estimation error}
We'll be very sketchy in this section.
\begin{thm}
Let $\wh f_{r,N,C}(x)$ be the least squares neural net estimator, calculated from $N$ samples $(x_i,y_i)$. Let the true function have $\ve{f}^*\le C$. 
Suppose that $\ve{\phi\pf{x}{\tau(\ep)}-1_{\ge0}}_{2}\le \ep$ for $\tau(\ep)$ polynomial in $\ep$. Then
\[
\text{MISE} =\E_{(x_i,y_i)} \ve{f-\wt f_{r,N,C}}^2 \le O\pf{C^2}{r} + O\pa{\fc{rp}{N}\ln N}.
\]
\end{thm}
The first ter, becomes smaller for $r$ large because more hidden nodes improves the best possible approximation. The second term becomes larger for $r$ large because the estimate degrades due to overfitting. Balance these terms by setting $n\sim \ve{f}^* \pf{N}{p\ln N}^{\rc 2}$; we get error $O(\ve{f}^*\pa{\fc p N \ln N}^{\rc 2})$. Note two things.
\begin{enumerate}
\item
The exponent $\rc2$ is independent of $d$.
\item Thus we can take $N=\wt O(kp)$ to get a fixed  approximation up to a fixed $\ep$.
\end{enumerate}•
In many other models (fitting to polynomials, etc.), the error goes like $\prc{N}^{\fc{C}{p}}$ or $\prc{N}^{\fc{2s}{2s+p}}$ for some smoothness parameter $s$. For fixed $\ep$, this requires the number of samples to be $\prc{\ep}^{\fc{p}{C}}$, exponential in the dimension.

Note, however, that there isn't a very good way to find the $\wh f$ other than brute-force trying all the parameters in an $\ep$-net, which is not an efficient algorithm.

\begin{proof}[Proof sketch]
\begin{enumerate}
\item
Let $\Te_{r,\ep,\tau,C}$ be an $\ep$-net for the parameters. Letting $\Te_{r,\tau,C}$ be the set of parameters with $|a_k|_1,|b_k|_1\le \tau, \sum|c_k|\le C$, we choose $\Te_{r,\ep,\tau,C}$ such that for every $\te$ there exists $\te^*\in \Te_{r,\ep,\tau,C}$ with $|a_k-a_k^*|,|b_k-b_k|^*\le \ep, \sum|c_k-c^*_k|\le C\ep, |c_0-c_0^*|\le C\ep$. We can $\Te_{r,\ep,\tau,C}$ having size
\[
\ln \Te_{r,\ep,\tau,C} \le \ub{(r(p+2)+1)}{\#\text{ parameters}} \ln\ub{ \pf{\poly(r)}{\ep}}{\text{grid width}}.
\]
\item
By a PAC-learning type bound (with least squares error rather than ERM---expected regret minimization for classification), we get
\[
\ve{f_{\text{best}}-\wt f} \le O(\fc{rp}{N} \ln N).
\]
Here the estimated parameters are
\[
\wt{\te}_{r,N} = \amin_{\te\in \Te_N} \pa{\rc N \sum_{i=1}^N (y_i-f_r(x_i,\te))^2}.
\]
\end{enumerate}
\end{proof}