\lecture{Wed. 2/9/2011}

\subsection{Functions of Bounded Variation}
\begin{thm}
A function can be written as the difference of two increasing functions if and only if it has bounded variation.
\end{thm}
\begin{proof}
Suppose that
\[
\Var(\psi;[a,b])<\infty.
\]
%dv
Then 
\begin{align*}
\psi(x)-\psi(a)&=V_+(\psi;[a,x])-V_-(\psi;[a,x])\\
\psi(x)&=[\psi(a)+V_+(\psi;[a,x])]-V_-(\psi;[a,x])%=\psi(x)-\psi(a)
\end{align*}
\end{proof}
\subsection{Convergence rate}
Let $f:[0,1]\to \R$ be smooth ($f\in C^1$)
Let
\[
R_n(f)=\rc{n} \sum_{m=1}^n f\pa{\frac mn}.
\]
\[
(R)\int_0^1 f(x)\,dx=\lim_{n\to \infty} R_n(f).
\]
How fast does $R_n(f)$ converge to the integral? In general we can't do better than the following argument:
\begin{align*}
(R)\int_0^1 f(x)\,dx- R_n(f)&=\sum_{m=1}^n \int_{\frac{m-1}{n}}^{\frac mn} \pa{f(x)-f\pa{\frac mn}}\\
&=\int_{\frac{m-1}{n}} \pa{x-\frac mn}f'(\xi_x)\,dx\\
&\leq n\cdot ||f'||_u\rc{n^2}=\rc{n}||f'||_u
\end{align*}
This is the best we can do because letting $f(x)=x$,
\[
R_n(f)=\rc n\sum_{m=1}^n \frac mn=\frac{n(n+1)}{2n^2}=\frac{1}{2}+\frac{1}{2n}.
\]

However, if $f$ is periodic we can do a lot better. We integrate by parts, in such a way so that we don't boundary terms, by choosing constants of integration appropriately. Then
\begin{align*}
&\quad (R)\int_0^1 f(x)\,dx-R_n(f)\\
&=-\sum_{m=1}^n (R)\int_{\frac{m-1}{n}}^{\frac mn} \pa{f(x)-f\pf{m-1}{n}}\,dx\\
&=-\sum_{m=1}^n (R)\int_{\frac{m-1}{n}}^{\frac mn} \pa{x-\frac{m-1}{n}}f'(x)\,dx\\
&=-\sum_{m=1}^n (R)\int_{\frac{m-1}{n}}^{\frac mn} \pa{x-\frac{m-1}{n}-\rc{2n}}f'(x)\,dx&\text{by periodicity}\\
&=-\sum_{m=1}^n (R)\int_{\frac{m-1}{n}}^{\frac mn} \pa{x-\frac{m-1}{n}-\rc{2n}}\pa{f'(x)-f\pa{\frac mn}}\,dx.
\end{align*}
We used periodicity to subtract off the average value of $x-\frac{m-1}{n}$ on $\ba{\frac{m-1}{n},{\frac mn}}$. (Note $\int_0^1cf'(x)\,dx=0$.) In the last step we used $\frac{1}{2n}$ is the average value of $x-\frac{m-1}{n}$ on the integral, so $\int_{\frac{m-1}{n}}^{\frac mn} \pa{x-\frac{m-1}{n}-\rc{2n}}=0$.
The last expression is bounded by $\rc{2n^2}||f''||_u$, better than $\rc{n}$.

Now assume that $f$ and $f'$ are both periodic $C^1$. (i.e. $f'(1)=f'(0)$ as well)
\begin{align*}
&\quad \sum_{m=1}^n\int_{\frac{m-1}{n}}^{\frac mn} \pa{x-\frac{m-1}{n}-\rc{2n}}\pa{f'(x)-f\pa{\frac mn}}\,dx\\
&=-\rc{2n}\sum_{m=1}^n \int_{\frac{m-1}{n}}^{\frac mn}f'(x)-f'\pa{\frac mn}+\sum_{m=1}^n \int_{\frac{m-1}{n}}^{\frac mn} \pa{x-\frac{m-1}{n}}\pa{f'(x)-f'\pa{\frac mn}}\\
&=-\rc{2n}\sum_{m=1}^n \int_{\frac{m-1}{n}}^{\frac mn}f'(x)-f'\pa{\frac mn}+\sum_{m=1}^n \int_{\frac{m-1}{n}}^{\frac mn} \pa{x-\frac{m-1}{n}}^2f''(x)
\end{align*}
Use periodicity to subtract off average value.
\[
-\rc{2n}\sum_{m=1}^n \int_{\frac{m-1}{n}}^{\frac mn}f'(x)-f'\pa{\frac mn}+\sum_{m=1}^n \int_{\frac{m-1}{n}}^{\frac mn} \ba{\pa{x-\frac{m-1}{n}}^2-\rc{3n^2}}\pa{f''(x)-f''\pa{\frac mn}}\,dx.
\]
Repeating this process. If $f$ and all its derivatives are periodic (of period 1) then the error in the Riemann approximation is going to 0 
faster than any $\rc{n^k}$.
%Restore to $f'$ factor $f'\pa{\frac mn}$.

To define Riemann integals for complex valued-functions, just look at real and complex parts separately. Let $f(x)=e^{2\pi ix}$.
Let $\xi_{m,n}=\frac mn\pa{1-\rc n}$. Then
\[
n\pa{\int_0^1 e^{2\pi i x}\,dx-
\rc{n}\sum_{m=1}^n f(\xi_{m,n})}=
\rc n\sum_{m=1}^n e^{\frac{2\pi i(1-\rc n)m}{n}}
=e^{\frac{2\pi i(1-\rc n)}{n}}\frac{1-e^{-\frac{2\pi i}{n}}}{1-e^{\frac{2\pi i(1-\rc n)}{n}}}\to -1
.\]
Moral: Periodicity destroyed by bad choice function.

\begin{df}
The \textbf{Bernoulli numbers} $b_l$ are inductively defined by
\[
b_{l+1}=\sum_{k=0}^l \frac{(-1)^k}{(k+2)!}b_{l-k}.
\]
\end{df}
\subsection{Lebesgue integration: Motivation}

Let $E$ be a set. Let $f:E\to \R$.
Let $\mu$ be ``volume" or measure of $\Ga$; that is $\mu$ is defined on nice subsets of $E$. We want to find a way to integrate $f$ with respect to $\mu$. We want to find a partition of $E$ into subsets $\Ga$ so that $f$ is constant or close to constant on each set $\Ga$. We then add up $\mu(E)y$ ($y$ this constant value).

Riemann: $E$ has topological structure and $f$ is nice with respect to topology (e.g. continuous). Partition into sets small from the topological standpoint, then give me $f$, it'll be nearly constant on each subset.

But if $f$ doesn't respect topology this FAILS!

Lebesgue: Look at sets of the form
\[
\{x|f(x)\in[(m-1)2^{-n},m2^{-n}]\}
\]
$f$ is nearly constant on each of these subsets, regardless of topological niceness. Now integrate.

But first we need to find the ``volume" or ``measure" of these sets! They will be HIDEOUS... Integration theory is easy compared to assigning measures.