\def\filepath{C:/Users/Owner/Dropbox/Math/templates}

\input{\filepath/packages_article.tex}
\input{\filepath/theorems_with_boxes.tex}
\input{\filepath/macros.tex}
\input{\filepath/formatting.tex}
%\input{\filepath/other.tex}

%\def\name{NAME}

%\input{\filepath/titlepage.tex}

\pagestyle{fancy}
%\addtolength{\headwidth}{\marginparsep} %these change header-rule width
%\addtolength{\headwidth}{\marginparwidth}
\lhead{Entire functions, Hadamard's Theorem}
\chead{} 
\rhead{} 
\lfoot{} 
\cfoot{\thepage} 
\rfoot{} 
\renewcommand{\headrulewidth}{.3pt} 
%\renewcommand{\footrulewidth}{.3pt}
\setlength\voffset{0in}
\setlength\textheight{648pt}

\begin{document}
\section{Jensen's Theorem}
\begin{thm}[Jensen's formula]\llabel{thm:jensenC}
Let $f$ be holomorphic in $B_R(0)$ with zeros $(z_k)_{k=1}^n$ (with multiplicity) inside. Then
\[
\ln |f(0)| = \sum \ln \pf{|z_k|}{R} + \rc{2\pi}\int_0^{2\pi} \ln |f(Re^{i\te})\,d\te.
\]
\end{thm}
%(Note this is true if some of the $z_k$ are on $C_R$ but requires a limiting argument.)
\begin{cor}[Jensen's inequality]\llabel{cor:jensenC}
Keep the setup. Then
\[
|f(0)|\ge \fc{R^n}{\prod_k |z_k|} \max_{|z|=R}f(z).
\]
\end{cor}
\begin{proof}
\begin{enumerate}
\item
Check that if $f,g$ satisfy the theorem then so does $fg$.
\item 
For $f$ nonvanishing, apply Cauchy on $\ln f$ (and take the real part).
\item
For $f=z-a_k$, we're reduced to showing $\oint \ln |e^{i\te}-a_k|\,d\te=0$. Interpret this as the mean value of a harmonic function, so it's 0.
\end{enumerate}
\end{proof}

\section{Hadamard's Theorem}

\begin{df}
The \textbf{order} of an entire function $f$ is 
\[
\inf \set{\al}{|f(z)|\precsim e^{|z|^{\al}}}.
\]
Let $z_k$ be the zeros of $f$ with multiplicity. The \textbf{genus} is 
\[
\inf \set{s}{\sum_k \rc{|z_k|^s}<\iy}.
\]
\end{df}

\begin{thm}\llabel{product-development}
Let $z_n$ be a sequence with $\lim_{n\to \iy} |z_n|=\iy$ (or a finite sequence).
%\item
%(Existence) There is an entire function with zeros exactly the $z_n$.
%\item
%(Uniqueness) All such $f$ are in the form
%\[
%f(z)=z^me^{g(z)} \prod_{n=1}^{\iy} \pa{1-\frac{z}{a_n}}e^{
%\frac{z}{a_n}+\rc2 \pf{z}{a_n}^2+\cdots +\rc{m_n}\pf{z}{a_m}^{m_n}
%}.
%\]
%\end{enumerate}
If $f$ is entire with order $\al<\iy$ with zeros $z_1,z_2,\ldots$ (with multiplicity, not including 0), then it has a product formula
\begin{equation}\llabel{product-formula}
f(z)=z^re^{g(z)} \prod_{n=1}^{\iy} \pa{1-\frac{z}{z_n}}e^{
\frac{z}{z_n}+\rc2 \pf{z}{z_n}^2+\cdots +\rc{k}\pf{z}{z_k}^{k}
},
\end{equation}
where
\begin{itemize}
\item $k=\fl{\al}$,
\item $r$ is the order of vanishing of $f$ at 0, and
\item $g$ is a polynomial of degree at most $a$.
\end{itemize}
The product converges uniformly locally. Moreover, 
\begin{equation}\llabel{num-zeros}
|\set{k}{z_k<R}|\precsim_{\ep} R^{\al+\ep}.
\end{equation}
Conversely, if $a=\fl{\al}$ and $z_k$ is a sequence satisfying~(\ref{num-zeros}), then the RHS of~(\ref{product-formula}) defines an entire function of order at most $\al$.
\end{thm}
\begin{proof}
(Following Stein-Shakarchi. Ahlfors seems shorter but I haven't read it.)
\begin{enumerate}
\item Massage Jensen to give us info about the count of zeros (cf. Abel summation). 

(1.2) Let $n(r)$ be the number of zeros in $B_r(0)$. Then $\int_0^R n(r)\fc{dr}{r}=\sum \ln \af{R}{z_k}$.

Proof. Just consier 1 zero, $z-z_k$. Now multiply.
\item (2) The relationship between order and genus. Let $\rh$ be the order. Then 
\begin{enumerate}
\item
$n(r)\le C_\ep r^{\rh_\ep}$ for some $C$, 
\item 
genus$\le \rh$.
\end{enumerate}

Proof. Plugging 1.2 into Jensen~\ref{thm:jensenC}, we get $\int_0^R n(r)\fc{dr}{r} \le C' R^{\rh'}$. Noting $n(r)$ is increasing, if it's $>C''R^{\rh'}$ infinitely often, summing $\int_R^{2R} C''R^{\rh-1}\,dr$ makes it $\succ R^{\rh'}$.
\item
The product is well-defined. (It is a natural guess; the exponential term is to make it converge.)
\begin{enumerate}
\item
(3.2) If $|F_n-1|\le c_n$, $\sum c_n<\iy$, then $\prod_{n=1}^{\iy}=F$ uniformly, $\fc{F'}{F}=\sum \fc{F_n'}{F_n}$.
\item (4.1) Let $P_k = z+\fc{z^2}2+\cdots +\fc{z^k}{k!}$.
Let $E_k = (1-z)e^{P_k(z)}$. Then for $|z|\le \rc2$, $|1-E_k|\le C|z|^{k+1}$, $C$ independent of $k$. 

Proof. 
\[
|1-E_m|=|1-(1-z)e^{-\ln (1-z)+O(z^{k+1})}|=O(z^{k+1}).
\]
\item The Hadamard product converges and has order $\le \al$.

Proof. Use the above criterion and 
\[
\sum (1-E_m\pf{z}{z_n}) \le \sum c\af{z}{|z_n|/|z|}^{k+1}
\]
finite if $\sum \rc{|z_n|^{k+1}}<\iy$. %Also this goes as $|z|^{k+1}$; the exponential goes as $e^{|z|^{k+1}}$.

To see it's order $\al$: terms for large $|z_n|$ don't contribute much; terms for small $|z_n|$ can contribute $\prod_{\af{z}{z_n}\le \rc2} (C|z| e^{C\pf{z}{z_n}^k})$. The $|z|$ contributes an $\ep$. In the exponent we have by Abel summation
\[
\sum_n C\af{z}{z_n}^k = \int^R z^k|z|^{\al-k+\ep} \ub{(\sum_{|z_n|<R} \rc{|z_n|^{s+\ep}})}{\le C}.
\]
Exponentiating gives the bound.
\end{enumerate}
\item
Given arbitrary $f$ of order $\rh_0\in [k, k+1)$, look at $\fc{f}{\text{Hadamard product}}$. It's entire and nonvanishing by construction; we need it to have order $\le \rh_0$ as well, so it equals $e^g$. Thus we need a lower bound on the growth of the Hadamard product.
\begin{enumerate}
\item 
(5.2) 
\[
E_k \ge\begin{cases}
e^{-c|z|^{k+1}},&|z|\le \rc2\\
|1-z|e^{-c'|z|^k},&|z|\ge \rc2.
\end{cases}
\]

Proof. 
For $|z|\le \rc2$, $E_k=e^{-\fc{z^{k+1}}{k+1}+\cdots}$; that first term dominates. For $|z|\ge \rc2$, $|1-z|e^{\bullet}=e^{-\cdots - \fc{z^k}{k}}$; the last term dominates. 
%The first part follows by (4.1). For the second part, 
\item (5.3) For any $\rh_0<s<k+1$, $\ab{\prod_{n=1}^{\iy} E_k\pf{z}{z_n}}>e^{-c|z|^s}$ when $z\nin \bigcup_{n}B_{|a_n|^{-k-1}}(a_n)$. 

Proof. Split into 3 products.
\[
\prod_{|z_n|\le 2|z|}\ab{1-\fc{z}{z_n}}e^{-c'\af{z}{z_n}^k} \prod_{|z_n|>2|z|} e^{-c\af{z}{z_n}^{k+1}}.
\]
\begin{enumerate}
\item
First: use the fact that $z$ is not in the balls.
\item Make $\prod e\to e^{\sum}$, separate out a factor $|z_n|^{-s}$.
\item Same idea.
\end{enumerate}
\item (5.4--5.5) We have $\bigcup B\subeq \set{z}{|z|\in I}$ where $I$ has finite measure. We can apply Cauchy to bound $\ln \pf{f}{\text{Hadamard}}$ on the valid radii, and get it's a poly of degree $\le s$.
\end{enumerate}
\end{enumerate}

\end{proof}

%\input{chapters/1.tex}
 
%\bibliographystyle{plain}
%\bibliography{refs}
\end{document}