\subsection{From independence to expansion and back again}

Find a $k$-independent function that can be efficiently computed (time/space).
%Time-space tradeoff for $k$-independent functions.

A data structure for representing a $k$-independent function $f:[u]\to [r]$ with evaluation time $t<k$ must use at least $ku^{\rc t}$ words of space.

Main result.
\begin{thm}
There is a randomized data structure for representing a $k$-independent function $f:[u]\to [r]$ with space $O(ku^{\rc t}t)$ and evaluation time $O(t\ln t)$. 
\end{thm}
Previous results: polynomials, graph powering, recursive tabulation.

Constructions of $k$-independent families of functions based on bipartite expander graphs. Neighbor function $\Ga:U\to V^d$.
\begin{df}
$k$-unique if for every $S\subeq U$ with $|S|\le k$ there exists $v\in V$ with exactly one neighbor in $S$.
\end{df}
\begin{lem}[Siegel]
$f(x)=\sum_i h(\Ga(x)_i)\pmod r$ defines $k$-independent family of functions.
\end{lem}

A $k$-unique $\Ga:U\to V^d$ gives a $k$-independent family. 

Most graphs give an optimal space-time tradeoff for $k$-independent hashing.

What's the problem?
\begin{enumerate}
\item
Explicit $k$-unique graphs with optimal parameters are not known.
\item
Storing a random $\Ga$ defeats the purpose of $k$-independent hashing (use less space!). 
\item
Verifying a given $\Ga$ is $k$-unique is infeasible.
\end{enumerate}•
Instead of sampling a random function, sample a function from a $k$-independent family. 
A $k$-independent function $\Ga:U\to V^d$ with $|U|=u_i$, $|V=O(ku^{\rc t}t)$ and $d=O(t)$ is $K$-unique with high probability.

Start with a $k$-unique function/graph. Increase the domain using a graph product. Define a $k$-independent family. Sample $\Ga\in \cal F$ and use to represent a neighbor function. Back to optimal parameters!

Randomized recursive construction of $k$-unique function. 
\[
\Ga(ax)=\bigopl_i h(a,\Ga(x)_i).
\]
If $\Ga$ is $k$-unique over $\Si^j$ and $\Ga$ is $k$-independent over $\Si^{j+1}$, then $\Ga$ is $k$-unique over $\Si^{j+1}$ whp.
Space $O(ku^{\rc t}t^2)$, time $O(t^2)$.

Divide and conquer. Recurse on each character.
\begin{df}
$k$-maj-unique if for every $|S|\le k$, exists $x\in S$ such that majority of vertices in $\Ga(\{x\})$ have unique neighbor in $S$.
\end{df}
If $\Ga$ is $k$-maj-unique, $\Ga(x_1x_2)_i=\Ga(x_1)_i\Ga(x_2)_i$ is $k$-unique over $\Si^2$.

We need to find $(x_1',x_2')$ with unique neighbor. 
Find $x_1'\ in \set{x_1}{(x_1,x_2)\in S}$ with $>\fc d2$ unique neighbors.
Find $x_2'\ in \set{x_2}{(x_1',x_2)\in S}$ with $>\fc d2$ unique neighbors.

View $x\in [u]$ as string of 2 chracters and recurse on each.
%Note $h$ must have $\ge k^2$

%rep theory and cs?
%random TM/circuit for indep?

Technique: graph products and alternating between expansion and indpendence. 

Open: Optimal expanders without $k$-independence?