\subsection{Super-resolution, extremal functions and the condition number of Vandermonde matrices}

Mathematical and algorithmic problem, connect to tools in analytic number theory.

Background and motivation from physics. 

There are fundamental limits to physics from optics: Rayleigh, Abbe limit. 

Can we recover find-grained structure from coarse-grained measurements?
2014 Nobel prize in chemsitry to super-resolution cameras, led to insights on biology.

Mathematical framework, Donoho, 1991. Consider a superposition of $k$ spikes, each $f_j$ in $[0,1)$. Signal
\[
x(t)=\sum_{j=1}^k u_j\de_{f_j}(t).
\]
Recover using coarse measurements. Integrating exponential. Get an evaluation of an exponential sum.
Only low freq: $|\om|\le m$.
\[
v_{\om} =\sum_{j=1}^k e^{i2\pi f_j\om}+\eta_\om
\]
Noise vs. noise-free setting.

When can we recover the coefficients and locations from low frequency measurements?

Prony's method, Pisarenk, Matrix pencil.
When no noice, there is a polynoimal time algorithm to recover $u_j,f_j$ with $m=2k+1$, at $\om=-k,\ldots, k$. 

What is possible in the noise-free vs. noisy setting is different!

Get an estimator $\wh{\om_j}\to u_j, \wt{f_j}\cdots $ which converges at a polynomial rate in $|\eta_\om|$.

There is a poly time algorithm for noisy super-resolution if $m>\rc{\De+1}$ separation condition. Let $d_w$ be wraparound distance in $\R/\Z$. $\De=\min_{i\ne j} d_w(f_i,f_j)$. %aliasing effects

\[
\min_\si\max_j|\wh{f_{\si(j)}-f_j}|+|u...|\le \ep
\]
poly!
drive to inverse poly level and get recovery.

Sharp threshold. As soon as $\fc{1-\ep}{\De}$. There is pair of separated exponentially close. Fejer kernel.

Asymptotic bounds for $m=\rc{\De}$ on grid.

Convex program for $m\ge \fc{2}{\De}$ with noise. 

Noise-free case: What would make it noise-stable? Let $\al_j=e^{2\pi i f_k}$, $V_m^k$ be the Vandermonde problems. It plays a key role in exact inverse problems; poly interpolation, sparse recover, inverse moment problems...

We create 2 matrices and set up generalized eigenvalue problem. $A=VD_u V^H$. Entries of $A$ correspond to measurements with $\om\in [-m+1,m]$. Similarly, companion matrix $B=VD_uD_\al V^H$. Similarly for $B$. 

Generalized eigenvalue problem gives oslution. The only things that work are $\la=\rc{\al_j}$.

Noise stability?

Vandermonde matrix has full column rank iff $\al_j$'s are distinve. This is enough for noise-free recovery. When is matrix pencil method robust to noise? When is generalized eigenvalue problem. Steward, Sun: various stability bounds for generlized eigenvalues/vectors based on condition number.

Show phase-transition for condiiton number of Vandermonde matrix.

Use extremal functions to bound the condition number of Vandermonde matrix. Used to prove sharp inequalities on exponential sums in analytic number theory.
\[
\ve{V_m^ku}^2 = (m-1\pm \rc{\De}) \ve{u}^2.
\]
Tight by Fejer kernel. Smallest SV is exponentially small.

Beurling-Selberg majorant. A function always above sign function and smooth, closest.
\begin{enumerate}
\item
$\sign(\om)\le B(\om)$.
\item
$\wh B(x)$ supported in $[-1,1]$.
\item
$\int_{-\iy}^{\iy} B(\om)-\sign(\om)\,d\om = 1$
\end{enumerate}•
inverse theorem in F analysis
\[
\pf{\sign(\pi \om)}{\pi}^2\pa{\cdot}
\]
Count primes in some interval: majorize and minorize interval and lower and upper bound how fall inside. Once have majorizer ad minorizer, easy.

Take a step back. 
Many inverse problems are well-studied in the exact case.
When is the solution robust to noise?
\begin{enumerate}
\item
polynomial interpolation: Lagrange interpolation. 
$[p_i]V=[p(\al_i)]$: find 

Highly unstable, often need exp small error in degree.
Statement about reals!

Ex. roots of unity, FT,well-conditioned.
\item
Sums of exponentials: Moitra and Saks. become well-posed in complex plane. If known $f(x)\pm$noise in circle, promise $f\le 1$ in larger: Hadamard 3 circle!

Noisy inverse problems are better posed over complex plane.

Connections between test functions in harmonic analysis and preconditionrs.

Give a way to obliviously rescale rows of unknown Vandermonde.
\end{enumerate}•