\subsection{Online matrix multiplication}

Our new OMv conjecture is a good choice for proving condiional lower bounds on dynamic problems

\subsubsection{Dynamic problems}

Example: dynamic single source reachability (ss-Reach).
\begin{enumerate}
\item
Preprocess a directed graph.
\item
Give a sequence of updates (insert/delete edges) and queries (can you reach node $n$ from start.
\end{enumerate}
Two types of update time amortized lower bounds imply worst-case loewr bounds.

The naitve algorithm is to grow the BFS tree from $s$ after each update. The update time is $m$, worst-case. Can we improve the update time from $m$ to $m^{1-\ep}$?  

The only unconditional lower bounds give polylog time. So people use conditional lower bounds: assuming conjecture $X$, get polynomial lower bounds.

There are many popular conjectures:
\begin{enumerate}
\item
Boolean matrix multiplication
%no $O(n^{3-\ep})$ combo algorithm
\item
3SUM
\item Multiphase
\item Triangle
\item APSP
\item SETH
\end{enumerate}
BMM, multiphase, 3SUM, and Triangle give hardness for SS-Reach. They give: you can't have better update/query time simultaneously. Each has drawbacks, because they only address combinatorial algorithms, are worst-case, not tight, etc.

Before our work, there are many conjectures, each with drawbacks.

\subsubsection{OMv conjecture}
\begin{conj}[OMv conjecture]
Given boolen $M$, output $Mv_1,\ldots, Mv_n$ online. There is no $O(n^{3-\ep})$ algorithm.
\end{conj}
BMM conjecture implies (in some sense) OMv conjecture. Fast matrix multiplication does not work in an online setting.

There are fast $\wt O(n^2)$ for special matrices. General algorithm best is $O\pf{n^3}{(\ln n)^2}$.
Unconditional lower bounds in related settings: algebraic circuit lower bounds, cell probe model.

Implications.
Using OMv we get sronger, amortized lower bounds (many tight), and simpler reductions. 
%Tight means that if you improve $X$, you worsen $Y$.

OMv and SETH are enough to prove all known strongest amortized lower bounds for dynamic problems.

Back to our first question: Can we improve update time $m$ to $m^{1-\ep}$.

\subsubsection{Proving hardness using OMv}
We first reduce to an intermediate problem: the edge query problem. Ask if there is an edge between 2 subsets $L,R$ on either side of a bipartite graph.

\begin{thm}
If there is an algorithm for edge query with preprocessing $\poly(n)$ time and $n^{3-\ep}$ time, then refute OMv.
\end{thm}
We reduce SS-Reach ($n^{1-\ep}$ update, query $n^{2-\ep}$) to edge query.

Make directed graph for ss-reach mathching the edge query graph. Update the edges out of $s$, into $t$, with the set, using $n$ queries. A fast algorithm for sS-reach gets a fast algorithm for edge queriy. $O(n^2)n^{1-\ep}+nn^{2-\ep}=O(n^{3-\ep})$.

\subsubsection{Summary}

All known conditional lower bounds for amortized update time fo dynamic problems; lower bounds from OMv is as high as from others. We get 15$+$ tight bounds. 

\subsubsection{Open problems}

Minimized spanning tree: still $\Om(n^.1)$ boudn for worst-case, ss-reach for decremental setting, maximum matching.