\subsection{Sum of Squares lower bounds from pairwise independence}

Constraint satisfaction problems: $n$ boolean variables, $m$ $k$-ary constraints.


Random assignment satisfies $\fc{|P^{-1}(1)|}{2^k}$ constraints. Can we do better? Yes for MAC-CUT, 2LIN, 3MAJ. In other cases, random assignment is optimal: $k-SAT^*$, etc. Approximation-resistant.

Understand as we gradually increase the number of satisfying assignments. 

Can beat random on most predicates with sparsity $O(k^2/\ln k)$.

%Most with $\Om(2^n/\sqrt k)$
%np-hard beat 
NP-hard to beat random for CSP(P) whenever $P$ supports pairwise indep subgroup: Exists prob dist unif on subgrp inside $\Pj^{-1}(1)\ge \rc 4$

What about what space in middle?

CSP(P) is pairwise independent if prob dist $\mu$ on $B^k$.

Captures most of white space. Most predicates with sparsity $\Om(k^2)$ are p.i. Assuming UGC, p.i. CSPs are approimation resistat.

$\Om(n)$ round Serali Adams and basic sdp cannot beat random.

SoS missing. 

Hierarchical strengthening of SDP.

SoS SDP highly successful convex relaxation for comb optimization
\begin{enumerate}
\item
sparsest cut
\item 
sparse coding
\item
attach on UG.
\end{enumerate}•
All knwn hard instances for weaker hier in poly time of UG, Bal sep, max cut.

SoS against CSP?

$k$-XOR's requires exponential SOS. Predicates supporting pairwise dindependent subgroup require sexponential SOS. 

\blu{
SOS needs exponential time to beat random for pairwise independent CSP's.
}

Previous proofs: reduction to resolution width lower bounds. algebraic structure crucial.

Our proof: local Gram Schmidt, does not depend on algebraic structure

At a high level, SoS SDP: 

CSPs: optimize over $B^n$. 

Solution: probability dist on $B^n$. 

Search over $e^n$ parameters.

Instead searhc over pseudo-distributions, satisfy only a subset of constraints of being a distributio!. Instead search over $n^{O(d)}$ parameters, $d=$degree/rounds.

Show instance $I$ with $M$ constraints. 
...
Deg $d=\ep n$ SOS thinks I is satisfiable.



Instance $I$ so that 
\begin{enumerate}
\item
val is $m||/2^k$. 
\item
$\wt E$ number I's constraints sat = m
\item 
pseudo-expe $\ge0$. 
\end{enumerate}•
New: We need to show (3).


New basis for small degree $\le d$ egree polys:
Satisfy positivity and pseudo-orthogonality wrt $\wt \E$ and $\wt [\chi_i^2]\ge 0$.

Expand a poly in this basis. %Observe 
\begin{enumerate}
\item
Start from Fourier characters in some order
\item
Run GS using $\wt E$ inner product.
\end{enumerate}
$\wt E$ is valid inner product restricted to any $d$ coordinates, corresponds to acutal expectation associated with a local prob dist.

But GS is highly sequential, long range dependent. 

Rely on: Expansion in the hyper graph $I$.

%All subsets of $\sim \ep n$ contraints cover $(k-1-o(1))/...$ vars

Define distance measure on subsets of size $\le d$.If $S,T$ are far enoug, thn the local distributions on $S,T$ given by $\wt E$ are independent.Orthogonalize only in a local neighborhood.

\begin{enumerate}
\item
fourier characters in carefully chosen order.
\item
Run step of GS using $\wt\E$ inner product.
\end{enumerate}•


Open: NP hardness. 





