
\subsection{Inapproximability of combo problems as LPs}

Encode a problem into a polytope $P$, $Q$ projecting to $P$. 


Problems: 

Say something about problem, not polytope. How you encode as polytope gives different ower bound.

Approx by scaling.


An optimizationproblem has set of feasibles, set of instances, oset of objective functions, $\cal S,\cal F, $val. Conider only maximization problems. 

Ex. Vertex cover.

Formulation complexity. $(C,S)$-approximable LP with realiation vectors, affine functions, guarantees $\max(Ax\le b)\le C(f)$.

\[
fc_+(\cal P, \cal C, \cal S)=\min\set{r}{\exists LP\text{ with }r\text{constraints capt }\cal P}.
\]

Factorization theorem. 
\begin{enumerate}
\item
$(C,S)$-approximate slack matrix $M$ of $\cal P$ is $M(f,s)=C(f)-\val_F(S)$. Factorization is $M=TU+\mu \one$, $T$ with $r$ cols, $\rank_{LP}M= $smallest r.
\item
$fc=\rank_{LP}M$.
\end{enumerate}

Optimal LP: Optimal LP's have same form. All info in problem encoded in... %objective fns

Objective functions point towards negative octant. Reconstruct LP from factorization of smallest size. 

Reduction mechanism: convert inapproximability results of problem to second problem. Given $P_1,P_2$,a reduction consissts of 
\begin{enumerate}
\item
map of feasible solutions $\ga :S_1\to \conv(S_2)$
\item
map of instances $\be:\cal F_1^{S_1}\to cone(\cal F_2^{S_2})+\R$.
\end{enumerate}•
val guarantees.

When there is a reduction, cna bound cc of $\cal P_2$ by $\cal P_1$. $\be$ only act on set of solutions, $\ga $ only on instances, independent. 

Hard gap between $C,S$. Measure hardness of gap: not possible LP with small number...

Factor soundness/completeness.

reduction from MaxCUT to vertex cover. 
\begin{enumerate}
\item
$\cal S$: all cuts ($B^V$)
\item
$\cal F$ subgraphs of $K_n$
\item
$\val$ objective functions $\val_H(s)=|\set{\{v_i,v_j\}\in E(H)}{s(v_i)\opl s(v_j)=1}|$.
\end{enumerate}•


...
fc is superpoly for approx guarantee of $\rc2+\ep$ for Max-CUT. SDP: ? conjectural.

All results independent of P vs. NP.

Open
\begin{enumerate}
\item
Hardness of SDP base problem
\item
Reduction for TSP: reductions aren't independent---solutions mapped according to instance.
\item Are there complete problems under this reduction?
\end{enumerate}•
