\subsection{Parallel repetition for general games}

2-prover 1-round gmes. 


$(x,y)\sim \mu$.
THey win if $V(x,y,a,b)$. 
$\val(G)=\max_{a,b}\Pj\pat{win}$.
$c$ is the answer length. $c=|a|+|b|$.

Can allow them to share randomness.

Ex. CHSH game. $x,y\in_RB\times B$. 
$V=1$ iff $a\opl b=x\wedge y$, value .75.

Projection game: for all $x,y,z$ there exists unique $b$< $V(x,y,a,b)=1$. Unique: for all $x,y,a$ $\exists b$ and vice versa.

Required to output $n$ answers. $a_i(x_1,\ldots, x_n)$, etc.
Win if $V(x_i,y_i,a_i,b_i)=1$ for all $i$.

Clear that $\val(G^n)\ge \val(G)^n$. Is the reverse true: is the best strategy to play independently on all coordinates? No, counterexamples of Raz08.

\begin{enumerate}
\item
Verb95: $\val(G^n)\to 0$ if $\val(G)<1$. 
\item
$\val(G^n) \le (1-\ep^{32})^{\Om(n/e)}$.
Improved to 3. 
%\item
%FV: $\exists \val(G),\fc 34, \val(G^n)\ge\rc8$
\item
For projection: $(1-\ep^2)^{\Om(n)}$. 
Raz08: unique game with squared dependence (connection to unique games conj?)
\end{enumerate}
PCP amplicfication: NP-hard to decide if $\val(G)$ or $\val(G)\le 0.99$. Parallel repetition: $\de$.

Connections: UGC, foam, ...

That was high-value regime.

Small value regime? DS13: 
\[
\val(G^n)\le \de^{\Om(n)}.
\]
Optimal NP-hardness for approximating set cover. 
%PCP of Mosh and Raz.

\begin{thm}
$\rc{2^c}\le \de \le \rc2$, 
$\val(G)=\de\implies \val(G^n)\le\de^{\Om\pf{n\lg\prc{\de}}{\ep}}$.

$\de\le \rc{2^c}$ transition (random guessing) $\de^{\Om(n)}$. 
%FV exmple.
\end{thm}

Reprove PRep theorems.

FIx a strategy for $G^n$. $W$ event of winning in all coordinates. If$\Pj(W)\ge$, design a strateg yfor G with $\Pj>\de$> 

Embed $(x,y)$ into random coordinate, sampel rest accoring to appropriate distribution and run for $G^n$.

Tries
\begin{enumerate}
\item
Set $X_i=x,Y_i=y$, uniform on others. Best boudn $\ge \Pj(W)$.
\item
Need into zoom into $W$. Sample $A_i,B_i|X_i=x,Y_i=y,W$ approximately. How to sample? Alice, bob only know $x,y$, resp.
\item
$X_i=x,Y_i=y$, jointly sample $R|X_i=x,Y_i=y,W$. (A bunch of properties) Alice and Bob should know $W$. (?) Also breaks the dependencies. Once sample correct $R$, can forget $W$. 
\end{enumerate}

Want mutual informations to be small. 

Exhibit $R$.

What if they know approximate distribution? How jointly sample? Use correlated sampling theorems. If approximately know some distribution. Some procedure sample approximately. 

What's $R$? Inspired by previous PR proofs and direct products in communication complexity BRWY13. 

?? Pic. Create various symmetries.


Open problems
\begin{enumerate}
\item
PR for general entangled games (pre-share entanglement), even for high value regime.

Holds for projection games, and product distributions.
\item
PR for 3 prover games, even high-value regime.
Define info complexity for multiparty.
\end{enumerate}•