\subsection{An interactive information odometer}


\[
IC_\mu(\Pi):=I(\Pi;X|Y)+
\]
what bob learns about $X$ and what ALice learns

%\]
%IC_\mu(f)=\inf
%\]

Information is amortized communication. 
\[
\lim_{n\to \iy} CC_\mu^\ep (f^n)/n = IC_\mu^\ep(f).
\]
Sharp threshold: $IC^{\fc 23}_\mu(f)=I$. WIth $\ll nI$ cc, success in computing $f^n$ is $2^{-\Om(n)}$. Like PR theorem for information complexity. Robust characterization.


More applications to CC.
\begin{enumerate}
\item
Information-theoretic secure communication between untrusted parties. 
%compile $\Pi$ into robust $\tau$. No matter what do, guarantee she never reveals. Assume no prior knowledge about honesty.

NO crypto assumptions!
\item
Interactive compression. Execute long protocol reveal few info, try to compress. Can be reduced to potentiall easier problem: IC$=\lg (CC)$.
\end{enumerate}

All related to information odometer problem. Same bottleneck. 

Maintain an online estimate of the IC of $\pi$ up to constant factor, with little information overhead.

Why does this task require interaction? Internal information cost is a function of both inputs. Ex. $\pi$: alice sends $x$ with probability 0.01, and $z\in_R B^n$ otherwise. Bob never knows whether he's learning something: marginal distribution is same. Variance is huge. Alice knows but not hard to adapt to get neither company can unilaterally estimate the quantity.

Can break $\pi$ into individual bits send. 

If Alice is the speaker in round $i$, $M_i\sim B(p_i)$.

$p_i=\Pj(M_i=1|xm_{<i})$, $q_i$ for bob.

Bob's prior $B(q_i)$ Bob learns from $M_i$: $D(p_i||q_i)=\Te(p_i-q_i)^2$ when bounded away from 1.

keep a constant factor estimte for all paths $m$ for all $t$.
%clicks once very 5 bits revealed. 
Clicks with prob $(p_i-q_i)^2$ prob each round. 

Primitive problem: goal: output 1 w.p. $(p_i-q_i)^2$ revealing $\precsim (p_i-q_i)^2$ btis of info.

Naive attempts.
\begin{enumerate}
\item
Send $p_i$ explicitly.
\item
Correlated sampling.
%entropy of Stat distance. The squared distance is lower.
\end{enumerate}
Main result: There is 2-round protocol that is
\begin{enumerate}
\item
correct: $\Pj(\tau\text{ outputs }1)=2(p-q)^2$.
\item
low info: $IC(\tau)\le O(H(p-q)^2)$.
\end{enumerate}•
Running with prob $\approx \rc{I}$ each round. Number of clicks $\approx 2I_{xy}(m_{<t})$, information overhead $\le O(\lg C)$.

Alice samples $Z_p\in [0,1]$ from density function $\mu_p(z)$. Alice sends $Z_p$ to Bob, Alice sends $(Z_p>p)$, Bob sends $(Z_p>q)$. Click (output 1) if inconsistent.

Correctness: area is $2(p-q)^2$. 

Main lemma: $D(Z_p||Z_q)\le 8(p-q)^2$.
%log ratio between densities cancels out.
%divergence is comparable only to 2nd 
%similar to Raz's PR. inspired by his distribution.
%exchange I^q, I^p. $H(click)\le H(2(p-q)^2)$.
Data processing, bounded by entropy of clicking. 

ex. if Alice sends a random string, prob click is 0. If $E=1$, for all $i$ get $>0$.

Abort if number of clicks $\ge 10$.

With small variation, can implement odometer even in adversarial setup (truthful mechanism for solicitng information). 

%more than 2 parties?