\subsection{Inapproximability of Nash equilibrium}

A short algorithmic history of Nash in 2-player games: 
\begin{enumerate}
\item
Finding a Nash equilibrium in 2-player games is PPAD-complete.
Finding $n^{-\ep}$-Nash equilibrium is still PPAD-complete.
\end{enumerate}
What about constant? There is a quasi-polynmial algorithm for finding $\ep$-Nash with constant number of players. Can we do better than $n^{\ln n}$?

Our focus is multiplayer games.
Fact: normal form representation is exponential in number of players. 
Instead, we consider oracle access to the payoff tensor.

Exponential lower bounds for this model!

Talk about restricted classes of games. Succinct representation.
\begin{enumerate}
\item
input is poly-size circuit implementation of value-query oracle
\item
each pair of players simulataneously palys a seaparate two-player game. Same strategy in each 2-player subgame, utility is sum
\item
Utility depends only on actions of small number of neighbors.
\end{enumerate}
\begin{thm}
Degree 3, bipartite, polymatrix game where each player has 2 actions: $\ep$-approximate Nash is PPAD-complete.
\end{thm}

How do we know right restriction to make? Get nice implications!

This is the ``3SAT of PPAD."
If you don't see anything to reduce from, use 3SAT. Similarly here.

Def. Arithmetic circuit. Generalized circuit: allowed to close the cycle. This defines CSP. 
$\ep$-GCircuit: find assignment $x:V\to [0,1]$ such that for all gates $(G,v_1,v_2,v)$, $x[v]=f_G(\pm \ep)\pm \ep$.

Not sure who Alice is playing with. %yang cai

$\ep$-approx Bayesian Nash equilibrium

%Eahc polywer playing ptimally

Two player game, constant number of actions, $\ep$-approx BN equilibrium PPAD-complete. (only number of types large)

All for additive approximations. Can talk about relative approximations. 

$\ep$-well supported relative nash equilibrium $\ge (1-\ep)$ of optimal utility. Das13: utilities in $[-1,1]$ is PPAD-complete (negative payoffs not standard). 

We get $[0,1]$.

Non-monotone markets. %CPY13
Raising the price of good $G$ while fixing all other prices strictly increases the demand of $G$.

Weak gross substitutibility.
Raising price of $G$ does not decrease demand of other items. 
For WGS, approx market equilibrium can be found in polytime.

PPAD-hardness for non-monotone markets: find $\rc n$-approximate market equilibrium is PPAD-hard. 
%Us: if utility of eahc is linear or from $U

Competitive equilibrium from equal incomes: equilibrium computed in practice, assign students to classes at Wharton. Stronger PPAD hardness for this problem. 

HPV construction. Average gadgets (DGP)

Cor: $\ep$-GCircuit, TW$\ep$RN, B, NMM, A-CEEI (course-match)

Beyond $\ep$-Nash

What is the PCP theorem for PPAD?

Natural equivalent: only satisfy $(1-\de)$ fraction of gates. Can be verified by reading a constant number of btis. Cojecture: $(\ep,\de)$-GCircuit is PPAD-complete. (equivalent of PCP theorem)

