\subsection{Fully homomorphic signatures from standard lattices}

Publicly verifiable computable (PVC): game between Alice, cloud server, and Bob. Alice uploads data $x$.
Arbitrary query/program $P$ to get answer $y=P(x)$. Bob has a ``pay-per-bit" channel: he wants to verify $y=P(x)$ without having seen the database.

Homomorphic signatures %intro in network coding.
Alice run signing agorithm wrt a secret key. Upload with corresponding signature to cloud server. Run to get result $y$. Can also run homomorphically over. $\si_{P,y}=eval_{pk}(P,x,\si)$. Bob downloads $y,P,\si_{P,y}$ (homomorphically derived signature)

\begin{enumerate}
\item
succinctness: only depends on output of $y$, not on runtime of $P$.

$preprocess_{pk}(P)=pk_P$. $verify(pk_P,y,\si_{P,y})=1$. (soundness)
\item efficiency: preprocess independent of runtime of $P$ and size of $x$.

A can sign many databases and evaluate many programs.
\item security: if $y=P(x)$, the cloud cannot convince that result is $y'\ne y$.

Challenger gives  apk to the adversary. Gives a data base $x$. Challenger gives $\si$. Eventually the adversary claims to have come up with $P,\si^*,y^*$. Adversary wins if valid signature of wrong result:
\[
verify(pk_{P'},y',\si')=1,y'\ne P(x).
\]
Relaxation: $x$ declared before pk (selective security). (Can also extend to many datasets)
\end{enumerate}

Constructions of homomorphic sigatures. 
\begin{enumerate}
\item
linear functions (bilinaear, RSA, SIS)
\item
bounded degree polys (ideal SIS , random oracle)
\item
(multilinear maps)
\end{enumerate}•

There is secure hom signature scheme for many databases and arbitrary programs represented by circuits where 
\begin{enumerate}
\item
succinctness: $\poly(\la,d)$, $\la$ security param, $d$ depth for $P$.
\item
efficiency: runtime of veri is same
\item
security assuming hardness of small integer solutions. 
\item context hiding: $\si_{P,y}$ reveals nothing about x.
\item locally computale
\item composable
\end{enumerate}

Compare to other solutions.
\begin{enumerate}
\item
CS proofs/SNARKs: cloud gives short proof verify$(y,n)=1$. All of them use non-standard assumptions which are essential.
\item
memory delegation: Bob needs multi-round (interactive) verification.
\item
etc. 
\end{enumerate}

How to construct? Homomorphic trapdoor functions, to homomorphic signatures. How to instantiate from lattices.

%selective to full, 1 to many

HTF is a family of functions parametrized by $f_{pk,x}:U\to V$. 
\begin{enumerate}
\item
given secret key sk, inverting $f_{pk,i}()$ is easy
\item claw-freeness: hard to find $f_{pk,0}(u)=f_{pk,1}(u^*)$. 
\item
Given $f_{pk,x_i}(u_i)=v_i$, exist deterministic algorithms 
$eval_{in}(C,(u_i,x_i)_i) \to u'$ and $eval_{out}(C,(v_i)_i)\to v'$.

Then $f_{pk, C(x)}(u')=v'$. 

Decouple comp over inputs and outputs separately, such that when apply to input get output.

How construct homomorphic signatures from HTDF? Fix public key, sample random from output space. 

Sample preimage corresponding to bits. 
$sign_{sk}(x_i)_i \to \si$: sample $u_i,f_{pk,x_i}(u_i)=v_i$. Run homomorphic... work over inputs. Homo derived signature. 
Run separate and evaluate...

$verify_{pk}(pk_C,y,\si_{C,y}) \to 1$ iff $f_{pk,y}(\si_{C,y})=pk_C$> follows by correctness of HTDF if $y=C(x)$.

If claw-freeness then secure. A gives $x$ to $C$. Sample signatures ourselves: $v_i=f_{pk,x_i}(u_i)$. Public:$ (pk,(v_i)), \si=(u_i)_i$. Adversary claims $pk, v,\si..$ A wins if $ver=1$. A wins if verify on $y'$ same for some $y'\ne y$. Is clawfreeness!

SIS problem: given $\Z\in (\Z/q)^{mn}$, find nontrivial short $z\in Z^m$, $Az=0$. 

Lattice algorithms: rapgen. short pre-image.

Fix $pk=A$, $sk=T$.$G\in (\Z/q)^{nm}$ global. $f_{pk_x}(U)=AU-xG$.

Want: given $sk$, invert $f_{pk,i}$ is easy.
\end{enumerate}•
%Lattice, HDT, fully holomomophic signatures.