\subsection{Lower Bounds on the Size of Semidefinite Programming Relaxations}
Unconditional computational lower bounds for classical combiantorial optimization problems in a restricted but powerful model of computation: all possible linear and semidefinite relaxations.

Goes back to Yannakakis in the 1980's, looking at flawed P vs. NP proofs. Refute all flawed proofs at once.

There are connections to optimizaiton and convex geometry. Settles open questions about semidefinite lifts of polytopes and ranks.
Identify optimal approximation  algorithms in the model. 

For every constraint satisfaction problem, the 
SoS (Lasserre) hierarchy gives optimal algoritm in this model.

Derive lower bounds for general model from known countersxamples for the SoS model.

Math programming relaxations: powerful generl approach for approximating NP-hard optimization problems. 

Three flavors:
\begin{enumerate}
\item
Linear (LP) ((A)TSP)
\item
spectral (Cheeger bounds)
\item
semidefinite (SDP) (MAX-CUT, SPARSEST CUT)
\end{enumerate}

Semidefinite relaxations subsume both.

Intriguing connection to hardness reductions, e.g., the unique games conjectures: plausibly optimal polynomial time algorithm.

Need to formalize algorithms. Mathematical programming relaxations: powerful general approach for approximating NP-hard optimization problems. 

Formalizes intuitive notion of LP relaxations for problem. Engough structure for unconditional lower bounds. Extends to SDP relaxations (but LP lower bound techniques break down). Not surprising: for some problems, SDP relaxations are strictly stronger.

Test computational complexity conjectures: approximation/PCP: unique games, sliding scale. Averge-case.

We consider the example of MAX-CUT. We give a LP formulations of MAX-CUT. Maximize $f_G(x)=\sum_{ij\in E(G)} (x_i-x_j)^2/4$ over $x\in \{-1,1\}^n$.

Equivalently, maximize $\sum_{ij\in E(G)}(1-X_{ij})/2$ over the cut polytope, $CUT_n=$convex hull of $\set{xx^T}{x\in \{-1,1\}^n}$.
The numebr of facets is exponential, so there is no small direct LP formulation.

The general size $n^d$ LP formulation of MAX CUT: The polytope $P\subeq \R^{n^d}$ defined by $\le n^d$ linear inequalities that projects to $CUT_N$. 

Often there is exponential savings: $\ell_1$-norm unit ball,...

The cut polytope doesn't have a LP formulation of polynomial size. We want to show this unconditionally.

Known: exponential lower bounds for LP formulations of MAX-CUT. Approximation ration $>\rc2$ require superpolynomial size. But the best known MAX CUT algorithms are based on SDP, and have approximation ratios better that $\rc2$.

Can we show size lower bounds for SDP formulations of MAX CUT?

A general size $n^d$-SDP formulation of MAX-CUT: spectrahedron $P\subeq \R^{n^d}\times \R^{n^d}$ defined by intersecting some affine linear subspace with psd cone. 

We show an exponetnial lower bound $2^{\Om(n^{.1})}$. Sum of squares is optimal SDP algorithm for MAX CUT.

Upper-bound certificates: algorithm with appoximate guarantee must certify upper bounds on the objective function $f_G$.  We can characterize SDP algorithms by their certificates. 

Certificates of degree $d$ sum of squares SDP algorithm: certify $f\ge 0$ for $f:\{\om 1\}\to \R^n$ iff $f=\sum_i g_i^2$ with $\deg g_i\le d$. 
%approx ratio $\al$: cert $f_G\le c$ for some $c\le OPT_G/\al$.

SoS captures Goemans-Williamson MAX CUT .878 approx:
\[
OPT_G-.878f_G\sum_i g_i^2
\]
with $\deg g_i\le 1$.

Relate general certificates to SoS certificates. For every $n^d$-size SDP algorithm, can certify $f\ge 0$ iff there exists $P\succeq 0$, for all $x\in \{\pm1\}^n$, $f(x)=\tr PQ(x)$. For the degree $d$ sum of squares SDP algorithm, $Q(x)=x^{\ot d}(x^{\ot d})^T$. A general SDP $Q$ captured by degree $d$ sum of squres if $\deg\sqrt{Q}\le d$.

Can simulate general $n^d$ size SDP algorithm by degree $O(d)$ (low degree) sum-of-squares.

For every low-degree matrix valued function $F$< can't distiguish between $Q$ and $Q'$ To try to approximate some object, bound the  complexity of the approximator. %larger than complexity of test functions? 
Key technical challenge: $\deg\sqrt{Q'}\gg \deg Q'$ at the heart of sum of squares counterexamples.

Take an indirect approach. find the simplest SDP algorithm $Q'$ that satisfies $\an{F,Q}=\an{F,Q'}$. Measure of simplicity is quantum entropy. There is a closed-form solution $Q'(x)\approx e^{tF(x)}$ where $t$ is entropy defect of $Q$, at most $\ln n^d$. Then expand $\sqrt{Q'(x)}$ and truncate the power series. Shows degree of apprxomator controlled by $F$.

Can simulate general small SDP by low-degree SDP.

Open questions:
\begin{enumerate}
\item
approximation beyond CSP and relatives (rule out .999-approx for TSP by poly-size LP/SDP)
\item
strong quantitative lower bounds for approx: rule out .999-approx for MAX CUT by $2^{n^{O(1)}}$-size LP/SDP
\item
rule out $2^{n^{.999}}$-size SDP for exact MAX CUT.

\end{enumerate}•