\subsection{Locally testable codes}

Weakly/strongly locally testable code:
\begin{enumerate}
\item
$\poly\prc{\ep}$/$O(1)$ queries, 
\item
$w\in C\implies T^w(z)=1$.
\item

\end{enumerate}

Main parameter is code length (amount of redundancy).

Goldreich and Sudan: there exists a strong LTC with linear distance and nearly-linear length. 
Viderman showed there is one with quasi-linear length.

In a LDC, we want to output $x_i$.

The best known LDCs re only superpolynomial length. Katz and Trevisan showed any $q$-query LDC must be of length $\Om(k^{1+\rc{q-1}})$. Introduce relaxed-LDCs which allow  output $x_i$ or $\perp$, ``I don't know."
\begin{enumerate}
\item
if $\de$-cose to $C$, $\Pj(D^w(i)\in \{x_i,\perp\})\ge \fc 23$.
\item
For all $w$ that is $\de$-close to $C$ and for most $i\in [k]$, $D^w(i)\ne \perp$.
\end{enumerate}

There exist relaxed LDC's with nearly-linear length.


Our gol is to construct short codes that are both strong-LTCs and relaxed-LDCs.

Motivation: the promise required for successfully decoding can be verified by testing procedure. %kill2 birds with 1 stone.

\begin{thm}
Exists a binary linear code that is relaxed LDC and strong LTC with ...
\end{thm}

Idea: each codeword consists of 3 equal-length parts
\begin{enumerate}
\item
large distance
\item
easy to decode
\item
consistency mechanism (PCP of proximity)
\end{enumerate}

\begin{df}
Verifier is not allowed to read input. PCPPs are PCPs with query access to both proof and input.

\begin{itemize}
\item
$x\in \implies \exists \pi, V^{x,\pi}=1$
\item
$x$ $\ep$-far: $\Pj(=0)>\fc 23$.
\end{itemize}
\end{df}
Exist PCPPs for NP with nearly linear length and constant query complexity.

$\pi_i$ is PCPP that asserts the first part of codeword of $C$ is consistent with the $i$th bit of the second part.
%not guaranteed any distance.

This yields quadratic legnth. With further ideas, nearly linear.

Weak testability: Because there exist PCPPs with nearly-linear length, 
any linear code can be transformed to LTC at the cost of nearly linear blow-up in the length. 

Drawback: results in LTCs are inherently weak

Solution: strong canonical PCPPs. 
\begin{enumerate}
\item
$x\in S\implies \redd!\pi, V^{x,\pi}=1$
\item
Strong soundness: proximity oblivious and must reject any $(x',\pi')$ with probability poly in distance from $\{(x,\pi(x))\}_{x\in S}$. %weight by lengths.
\end{enumerate}•

Construct scPCPPs for specific statements we need with poly length. $C$ good linear code: eists scPCPP...

How to obtain strong testability?
\begin{enumerate}
\item
Naive way yields poly length codes.
\item Use alternative approach relying on tensor codes.
Use scPCPPs ony for short statements: even wth poly blow-up, length of each proof still sub-linear.

Local consistency to global consistency.
\end{enumerate}•
%strongest conjecture