\subsection{Depth 5 lower bounds for iterated matrix multiplication}

%size and depth

$
IMM_{m.d}
$ is $(1,1)$-entry of the product of $d$ matrices of size $n\times n$. The number of input variables is $N=d\cdot n^2$. A monomial in IMM corresponds to a path in a layered graph. (Two consecutive layers represent a matrix.) 

Can IMM be made highly parallel. i.e., efficiently computed by small depth circuit?

VP vs. VNP.  VNP coefficient of any given monomial efficiently computable. Are there hard polys in VNP require super-poly size circuit?

Fan-in bounds: $\Si^a\Pi^b\Si^c\Pi^d$.

Any $size(f)=\poly(N,d)$ implies $f$ computed by depth 4 $\Si\Pi\Si\Pi^{\sqrt d}$ circuit. 
If $f\in VP$, $f$ has $N^{O(\sqrt{d})}$ size depth 4 circuit. Loer bound $N^{\om(\sqrt d)}$ implies VP$\ne $VNP.

%NW
IMM lower boudn $N^{\Om(\sqrt d)}$. Depth reduction tight at 4.

Depth 5: better depth reduction at depth 5? $f\in VP$ implies size$(f)=\poly(N,d)$.

Our result: $N^{\Om(\sqrt d)}$ for IMM, circuit $\cdots \Si^{N^{\mu}}$, $0\le \mu <\rc2$.
Better depth reduction not possible at depth 5 if bottom fan-in is restricted.

Generic strategy: define a complexity measure $\Phi$ maps each polynomial to a number. 
Dimension of projected shifted partial derivatives (DPSP). $M$ is set of multilinear monomials of degree $k$, $S$ for degree $l$.
\[
DPSP_{k,l}(f) = \dim_{\F} \spn(proj(\set{\be \pl_\al f}{\al\in M,\be \in S})).
\]

Process: randomly select a set of variables $V$ according to some nice distribution. Set variables outside to be 0. 

A small circuit is simplified (small $\Phi$). The polynomial still remains complex: large $\Phi$.

Circuit decomposition lemma: depth 5 circuit: $C$ bottom fanin $N^{\mu\i (0,\rc2]}$
\[
C|_V=C'
\]
$C'$ is $\Si\Pi\Si\Pi^{O(\sqrt d)}$ and $DPSP_{k,l}(g)=0$ for any $k,l$.

Small upper bound on DPSP.

Expand bottom two layers of gates.

$Q_{ijr}$ has 3 type of monomials
\begin{enumerate}
\item
small, total degree $<2\sqrt d$: put into $C'$
\item
large, exists var with degree $\ge 3$: DPSP is 0. (put to $g$)
\item
Large, each variable has degree $\le2$, support $\ge \sqrt d$.
\end{enumerate}•
Union bound fails for $\mu\ge \rc4$. Instead, apply mild restriction, makes bottom fan-ins $\le N^{\rc 4}$. Apply desired full restriction: analyze as above.

non-Chernoff concentration bound.

dimension of matrix is $d^q\times d^q$ where $q$ is large constant depedendent on fain-in $p$.

Large upper bound on DPSP(IMM), small upper bound on DPSP$(C|_V)$

IMM not efficiently compuetd by bottom fan in restricted,

Prove lower bound for $\Si\cdots \Si $...