\subsection{Streaming interactive proofs}

Given $g$, replaced $g$ with mulilinear $\wt g$ over large field; it has good error-correcting properties.

View $\mx:B^{\lg n}\to B$. Let $|\F|\ge 4\lg n$, $\wt{\mx}$ be the multilinear extension. $V$ evaluates $\wt{\mx}(r)$ in streaming pass over $\mx$ with space $O(\lg n\lg\lg |\F|)$. Is a sketch of input. Write $\wt{\mx}(r)=\sum_j x_j\wt{\de_j}(r)$.

Sends line to prover. $\la$ specified with $\lg n$ elements. Prover sends back the polynomial restriction to line. Making a claim about every point on the line.
%compare to data stream.
If the prover sends any other polynomial, %any 2 $\lg n$ polys have to
is disagrees almost eveyrwhere.
Only know $r$ is somewhere along the line and not equal to $i$.

RangeCount protocol: fix $[n], R\subeq 2^{[n]}$.

Input: $x_i\in [n]$, range $R^*\subeq R$. Goal to output $|\set{i}{x_i\in R^*}|$. Reduce it to the index problem. Verifier creates a derived stream. Define over different data universe. Every time see new stream update, insert every single range that contains $x_i$. 

Wat frequency of $R^*$ in derived stream: answer with generalized INEX protocol. But $V$ requires $|R|$ time per stream update.

%input vector exp big.
Use a higher-degree extension.

Online interactive proofs (communication model).
$O(1)$-round SIPs can solve INDEX. 
2 new hierarchies of communication models
\begin{enumerate}
\item
OIP${}_+[k]$ can simulate all $k$-message SIPs. 
\item
OIP$[k]$ weaker, but can still simulate all known SIPs, and capture the fundamental way SIPs differ from IPs.
\end{enumerate}

$AM^{cc}$ [BFS86]. Alice and Bob want to ocmpute $f(x,y)$. Merlin helps but is untrusted. Random coins are sent to both parties. Merlin gives answer, then communicate in deterministic protocol.

Restricted private coin variant. Secret coinc not seen by Merlin. Alice sends a single message to Bob. Then Bb and Merlin engage in $k$-message interaction.

In $OIP[k]$: for Bob and Merlin's message to be independent. Bob and Merlin engage in $k$-message interaction in step 2, not 3.

$OIP[2]$ protocol of cost $O(\lg n\lg\lg n)$ for INDEX.Toss secret coins to determine eval point. Send $\la$. Alice sends $\wt{x}(r)$.
%only depends on index i.

$OIP^{[k]}$ class of functions solved by polylog cost $OIP[k]$ protocols, $AM^{cc}$ functions solved by polylog cost $AM^{cc}$ protocols. Containment with exonential separation:
\[
OIP^{[1]}\sub \cdots \sub OIP^{[\ge 4]}=AM^{cc}.
\]
%TM for IP: collapse to second level.
2-message OIP are equivalent to $R^{[2,B]}$: without Merlin: Bob sends single to Alice, Alice sends single response. $OIP^{[2]}$Lpretend Bob can send 1 message to Alice even though lets neither Bob nor Merlin talk to Alice. 4: both Bob and Merlin can talk to Alice: simulate all of $AM^{cc}$.

