\subsection{PIT for ROABP}

Read-once oblivious: any variable occurs in at most 1 layer, and in specified order.

Polytime whitebox test, quasi-poly blackbox test.

First poly-time whitebox test for sum of 2 ROABPs. (key: variable orders can be different)

Evaluation dimension or partial coefficient dimension (Nisan 1991). %Write as $\F[x_ 

Computed by ROABP of width $w$ and var order $x_1,\ldots, $ iff for all $i\in [n]$, $\dim\set{A_e}{e\in B^i}\le w$.
\begin{proof}
\begin{enumerate}
\item
Write $A=\sum_{j=1}^w P_jQ_j$ where $P_j\in \F[x_1,\ldots, x_i]$ and $Q_j\in \F[x_{i+1},\ldots, x_n]$. (Partial coefficients.)

\item
Small coefficient dimension implies small width. Do layer by layer.
\end{enumerate}
\end{proof}

Given $A,B$ computed by ROABPs in different variable orders.
\begin{enumerate}
\item
Find linear dependencies among partial coefficients of $A$.
\item
Verify same dependencies hold among corresponding partial coefficients of $B$.
\end{enumerate}
Reduces to zero testing for ROAMP of a larger width $(\le w(w+1))$.
Generalize to $c$ ROABPs with complexity $\poly(w^{2^c}n^c)$. 

Make blackbox with least basis isolation and low-support rank concentration.

QUestions:
\begin{enumerate}
\item
Bring down time complexity to $w^{O(c)}$? 
\item Depth 3 multilinear? 
\item Inspired from nonequivalence of 2 ROBP's Are there more connections?
\end{enumerate}

\subsection{Matrix rigidity}
Rigidity: $V$ is $(r,d)$-rigid if for any matrix $V'$ where $d(v_i,v_i')\le d$ we have rank $\rank(V')\ge r$.

If $V$ is $(\Om(n),n^\ep)$-rigid, the linear transformation that maps a vector $x$ to $Vx$ does not have an $O(n)$-size $O(\ln n)$-depth linear circuit.

Valiant needs $(\Om(n),n^\ep)$-rigid. With high probability, a random matrix is $(.99n,\Om(n))$-rigid.
The best explicit matrices are $(.99n,0)$-rigid. The best are $(r,\Om(\fc nr \ln \fc nr))$, $r\ge (\ln n)^2$.

Take a matrix of full rank; Zaranckiewicz's lemma shows after some changes there will be an untouched matrix. ``Untouched minor barrier."

Design matrices are candidates for rigidity over $\F_2$. Design matrix; every row has $n^{1-\ep}$ ones, supports of every two distinct rows intersect by $n^{1-2\ep}$.

Goal:  Establish $(n^{2\ep +\de},n^{1-2\ep})$-rigidity of matrices $V_{m=\rc\ep}$.

Any rigidity proof needs to exhibit a property $\pi$ that is
\begin{enumerate}
\item
satisfied by all low rank matrices
\item not satisfied by $V_m$ even after perturbations.
\end{enumerate}•

Our property is approximability: after a certain particular embedding into $\R^n$ its rows amit a nontrivial approximation by a low-dimension Euclidean space.

Embedding $\F_2^n\to \R^n$, $x\mapsto \fc x{\ve{x}_2}$.
\[
A_r(V)=\max_{W\subeq \R^n,\dim W=r} \min_{x\in V} \ve{\Pj_W(x)}^2.
\]

$A_r(V)$ is equivalent to Kolmogorov width of $V$.

Strategy:
\begin{enumerate}
\item
Show $A_r(V_m)$ is small.
\item
Show that $A_r(V_m)$ is robust under perturbation of rows of $V_m$.
\item
Show that for low-dimensional $\F_2$-linear spaces $L\subeq \F_2^n$, $A_r(L)$ is large.
\end{enumerate}
Imply that $V_m$ have high rank even after perturbations.

%Had easy  to approx because fairly strange embedding. Division: 
The last one is hardest!

(6.4, 5, 7 in CDSM)
\begin{enumerate}
\item
$r=o(n)\implies A_r(\F_2^n)\sim \fc rn$.
\item
$r=\om(n^{1-\ep})\implies A_r(V_m)\approx \fc v n$.
\item
$r=\om(n^{1-\ep})$: every row of $V_m'$ differ from corresp of $V_m$ in $O(n^{1-\ep})$: $\approx \fc rn$.
\end{enumerate}•

Conj. 6.8 implies $(n^{2\ep+\de},n^{1-2\ep})$-rigidit of matrices $V_m$.
%find subspace all large component
%natural proofs limitation to rigidity?
%research drift?

%...see IAS talk.

need better approx for $\F_2$-lin spaces.. weak for hi-dim approx.
use more property of $\F_2$-lin than trignular raank.
approx for $n$-subsets of $\F_2$ linear spaces rather than complete spaces?
%rank, number of changes, rows.
