\subsection{Identifying honest $EXP^{NP}$ oracle among many}


We have two oracles, one is honest, one is dishonest. 

\begin{enumerate}
\item
Notion of selector: there exists a selector iff we are able to remove short advice.
\item
Existence of selector for $EXP^{NP}$-complete languages.
\end{enumerate}

Instance checker for $f$ checks if given oracle correctly computes $f(x)$ in poly time.
\begin{enumerate}
\item
There are instance checkers for $P^{\#P}$, PSPACE, EXP-complete.
\item Must be in $NEXP\cap coNEXP$. Note $NEXP\subeq EXP^{NP}$.
\end{enumerate}

Probabilistic selector for SAT. Make queries: is $\psi$ satisfiable. The honest oracle gives correct answers; the dishonest oracle gives arbitrary answers.

If the oracles agree, then the answer is right. The essence of the task is to determine which is honest when they disagree.
\begin{df}
A \vocab{selector} S for $L$ is prob oracle machine such that if $A_0=L$ or $A_1=L$, then
\[
\Pj[S^{A_0,A_1}(x)=L(x)]\ge 0.99.
\]
For a deterministic selector, deterministic succeeding always.
\end{df}

Selector for $P^{NP}$-complete languages: lexicogrphically maximum satisfying assignment. Output $k$th bit of lexicographically max satisfying assigment of $\ph$.

Given $(\ph,k)$ and 2 oracles, make queries $(\ph,i)$ to the oracles. If $v_0=v_1$, output the $k$th bit. Else WLOG $v_0<v_1$. If $\ph(v_1)=1$, 1 is honest, else 0 is honest.

$EXP^{NP}$-complete language: succinctly described Boolean formula $\Phi:B^{2^n}\to B$ and $k$. Output $k$th bit of lexicographically maximum satisfying assignment. The proof strategy is the same.

\begin{enumerate}
\item
Is $V_0<V_1$? binary search and PIT.
\item
Is $V_1$ a satisfying assignment of $\Phi$? Similar to $MIP=NEXP$.
\end{enumerate}•

Instance checker vs. selector. If we have a instance checker, we can construct a selector. The task of selectors is strictly easier than instance checking.

\bal
\SAT\in P/\log &\implies \SAT\in P\\
EXP\in BPP/\log & \implies EXP\subeq BPP.
\end{align*}
This follows from the instance checkability of EXP-complete languages.

We can remove short advice when we have a selector.

If there is a selector $M$, we can remove 1-bit advice. The two oracles gives answer $M(q,0)$ and $M(q,1)$.

%Polynomially many 

\begin{thm}
For any paddable $L$, TFAE:
\begin{enumerate}
\item
$\exists$ deterministic selector
\item 
for any oracle $R$, $L\in P^R/\log$ implies $L\in P^R$
\end{enumerate}
\end{thm}

Even given many oracles, we can still identify an honest oracle. Idea: tournament. Depending on their answer, divide them into 2 teams. Pick 2 and run, eliminate those we doubt.
%how many oracles?

%To remove advice of size $O(\lg n)$,
If no selector exists, then for some relativized world, even 1-bit advice cannot be removed. Construct the oracle by diagonalization.

Other results:
\begin{enumerate}
\item
There exists a deterministic selector for PSPACE-complete languages.
\item Any languages with deterministic selector in PSPACE.
\end{enumerate}•
Upper bound for probabilistic selector: $S_2^{exp}$.

Close the gap between $EXP^{NP}$ and $S_2^{exp}$. 

%Set of languages with select

Poly advice?





