\subsection{Quantum Arthur-Merlin}

Definition of IP, AM.
%$SZK\subeq AM\cap coAM$.

Quantum interactive proof system, introduced by Watrous. The verifier and prover use quantum computers, and communication is quantum.

\begin{itemize}
\item
$PSPACE\subeq QIP[3]$.
\item
$QIP=QIP[3]$. Contrast with classical IPs which conjecturally can't be parallelized into consant-turn IPs.
\item
$QIP=PSPACE$.
\item
$QIP[1]=QMA$.
\end{itemize}
Little is known about $QIP[2]$. 

QAM: Arthur sends a classical random string $y$; Merline sends a quantum state. 
$QMAM=PSPACE$.

Investigate 2-turn QIP systems more finely. What is a fully quantum Arthur-Merlin proofs system?

\begin{df}
qq-QAM (in between QAM and $QIP[2]$): Arthur creates polynomially many copies of EPR pair, where the first half of each copy is in S1, the scond half is in S2. He sends S2.

Merlin returns a quantum state $\rh$. Arthur decides to accept/reject from $x,\rh,S1$ by quantum algorithm.
\end{df}

qq-QAM has a natural complete problem CITM.
\begin{df}
Instance: quantum circuit $C$ which has some specified input qubits and $\ell$ specified output qubits.

Yes: there exists a state $\rh$ such that $D(C(\rh),\prc2^{\ot l})\le a$.

No: for any state, $...ge b$.
\end{df}
\begin{enumerate}
\item
do there exist two inputs which have approximately equal images (under 2 circuits)? QIP-complete.
\item
do two given pure states have ... QSZK-complete.
\end{enumerate}

\begin{thm}[Quantum analogue of Babai's collapse theorem]
qq-QAM does not change by adding $O(1)$ turns of classical interactins prior to the communications of the qq-QAM proof system.
\end{thm}

$t_m\cdots t_1-QAM(m)$ into 4 classes.
\begin{enumerate}
\item
PSPACE$=qcq-QAM=QMAM$
\item
cq-QAM
\item 
qq-QAM
\item 
qc-QAM
\end{enumerate}
\[
AM\subeq cc-QAM\subeq cq-QAM=QAM\subeq qq-QAM\subeq QIP[2]\subeq PSPACE.
\]

\begin{enumerate}
\item
New upperbound for QAM: $QAM=cq-QAM\subeq qq-QAM_1$. (cf. $QAM\subeq BP\cdot PP$).

QAM$\subeq QIP[2]_1$.
\item
$cc-QAM=cc-QAM_1$.
\end{enumerate}

$c\cdots cqqAM(m)=qq-QAM.$
Proof strategy:
\begin{enumerate}
\item
$m\ge 4$: Use Babai's classical proof.
%Verifier sends classical bits.
%$cccqq-QAM\subeq ccqq-QAM$. Error probabliy can be reduced... 
\item
$cqq-QAM\subeq qq-QAM$: use CITM.

Let $A=(A_{yes},A_{no})$ in $cqq-QAM$. On input $(x,q$), simulate the last 2 turns under the condition that the 1st message in $\Pi$ was $w$.

Show $qq-QAM$ completeness of MaxOutEnt: is the entropy of a  given channel large for any input?
\item
$ccqq-QAM\subeq qq-QAM$.
\end{enumerate}

\begin{itemize}
\item
CITM: is the output of a given circuit close to totally mixed state for any input?
\item
MaxOutQEA: does a quantum channel have the max output entropy larger than a threshod?
\end{itemize}•

Open:
\begin{enumerate}
\item
Natural problem in qq-QAM not known to be in cq-QAM. Or equl?
\item
Non-trivial loewr and upper bound for qq-QAM? $cq-QAM, QIP[2]$
\item
$qq-QAM=qq-QMA_1$.
\item Quantum analogue of Goldwasser-Sipser
\end{enumerate}
