\subsection{Algorithms for strategic agents}

What is the revenue-optimal auction for selling multiple items?

Myerson designed a revenue-optimal auction for a single item. Bids are transformed into virtual bids. The item is awarded to the highest virtual bidder.

People tried to extend it to more than 1 item. There are lots of characterizations for special settings. CS has contributed approximation algorithms. The settings are limited; there is no unified approach.

There are formal barriers to %strong characterizations and 
finding optimal auctions. %computationall y hard to find.
Challenges arise for 1 buyer and 2 items.

Treat as algorithmic problem.

Adding the second item, each bidder makes a bid on each item. Transform to virtual bid. Each item is awarded to the highest virtual bids.

Myerson's transformations have a closed form. Ours is randomized and computed by a linear program.

Put into context of algorithmic mechanism design. It's not traditional algorithm design. Input, algorithm, output.

We have to first learn the input from agents' report, and the output feeds the agents' payoff. They can lie. We call it a mechanism instead.

They motivated the following design. How much more difficutlt are optimization problems on strategic input comprared to honest input? We want a mechanism that works on strategic input. We want to put a black box around an algorithm that works on honese input, and get what we want by probing it.

Maybe we can reduce hard problems in algorithic mechanism design to hard problems in algorithm design. This allows a much larger community to work on these problems.

A reduction exists! We can also learn what makes mechanisms so challenging.

We can view seminal results as black-box reductions.

VCG is a black-box reduction: if you have the optimal algorithm for welfare, you can get the optimal mechanism for welfare. 
Agents report types. Input into an algorithm that works on honest input. Just choose the agent that's outputted.. There exists a payment scheme that makes this truthful. More recent works tries to make it approximation-preserving.

Probe multiple inputs instead.

Put virtual bids into the black box maximizing welfare.

We do care about objectives beyond welfare and revenue. Job scheduling on unrelated machines is paradigmatic. Machine $i$ takes time $t_{ij}$ to process $j$. Minimize makespan. Machines are strategic agents, you don't know how long it takes them to process the jobs.
We know how to truthfully optimize welfare (revenue) if we know how to algorithmically optimize welfare (virtual welfare). There is no reduction from truthfully minimizing makespan to minimizing makespan even in 1 dimensions. A reduction exists if we perturb algorithmic objective. 

\begin{thm}
There is a polynomial time reduction from mechanism design for objective $O$ to algorithm design for same objective $O$ plus virtual welfare.
\end{thm}
The reduction is approximation-preserving; there is no restriction on types (dimensions).

``Transitioning from honeset to strategic input is no more computationally difficult than adding virtual welfare to objective."

For now, objective is revenue, and agents are additive. (The value for a set of items is the sum of the values.)

A bidder's type is a vector of values. They are taken from some distribution. (Beliefs are modeled as this distribution.) Bayesian incentive compatibility: if every other agent is telling the truth, I want to tell the truth as well.

A mechanism is a function that takes a input a profile of types and outputs a distribution over allocations and a price for all agents. We can think of a mechnsm as a large-dimensional vector (for every input, what's the output). Thinking of it this way we can try to optimize over all algorithms. 

We can write out linear constraints to make $M$ truthful and makes it a feasible mechanism.

The problem is that the input is enormous (too many variables). We need a compact description.

Instead we use the reduced form. We just keep the important information.
\begin{enumerate}
\item
For every agent and item, what's the probability they get the item?
\item
What's the expected price they pay?
\end{enumerate}
Write a LP with much fewer variables. We can still write linear constraints. The new challenge is to write constraints to guarantee it's feasible, correspond to an actual algorithm. This is challenging. Does a reduced form correspond to auction?

There is a different succinct description that we can use, similar to reduced forms.

Given a reduced form does it correspond to actual mechanism? Borders theorem gives conditions for feasibility. Borders requires checking exponentially many inequalities. We give equivalent conditions that can be efficiently checked.

We can solved LP using computationally efficient Border's Theorem. For general settings, use equivalence of separation and optimization.

Every reduced form that is a corner of the space of feasible reduced forms is a virtual welfare maximizer for some virtual transformation. The direction has a component/weight for each $i,j,l_i$. Optimizising in this direction is like optimizing virtual welfare.
Combine with Caratheodory's theorem: every point in $P$ is a convex combinations of corners of $P$. The optimal mechanism is a distribution over virual welfare maximizers. 

Sample corners appearing in convex combination.

\begin{enumerate}
\item
Write succince LP for optimal truthful mechanism using reduced form
\item Solve LP.
\item
\end{enumerate}

Contribution to pure convex optimization: extend to accommodate approximation errors. We apply it to new auction settings where auctions have budgets.

No matter what the bidders' types, the revenue poptimal auction is a distribution over cirutal welfare maximizers. We discovered the structure through algorithmic study.

How much more difficult are optimization problems on strategic input comapred to honest input? For all objectives in Bayesian setting, reduction exsits from mechanism to algorithm design with perturbed objective.

Look towards future: %dominant strategies 
New framework and tools for multi-dimensional settings. What can we do with them?

%structure deal with 
%optimize over another function
%Whatever optimum is subject to truthfulness, get for that.
%hopeful for theory/experiments. 