%coresets

\section{Parallel computation}
\subsection{Randomized composable core-sets for distributed optimization, Vehab Mirrokni}

%diversity maximization
%randomized clustering

Large-scale graph mining
%\begin{enumerate}
%\item
Algorithms, tools: Ranking, pairwise similarity, graph clustering, balanced partitioning, embedding.
%\item
%\end{enumerate}•

Problems
\begin{enumerate}
\item
diversity maximization: composable core-sets
\item
clustering problems: mapping core-sets
\item
submodular.coverage maximization: randomized composable core-sets.
\end{enumerate}•

Composable core-sets are useful. 
Inspired by processing big data. Extract and process a compact representation of data:
\begin{enumerate}
\item
sampling: small subset
\item
sketching
\item
\end{enumerate}

Divide the input of size $N$ into pieces $T_1,\ldots, T_m$, and send sets $S_i$ back. Run the algorithm on the $S_1,\ldots, S_m$ to find the output set.

Key example in combinatorial geometry. Ex. farthese distance pair of points. An $\al$-core set is $S$ with $f(S)>\al f(P)$.
\begin{enumerate}
\item
The convex hull is a 1-core set.
\end{enumerate}
Composable core-sets: partition input into parts $T_1,\ldots, T_m$. In each part, select $S_i\subeq T_i$. Let $S=\bigcup_i S_i$.

$f_k(S):=\max_{S'\subeq S,|S'|\le k} f(S')$. ALG(T) is output of ALG on input set $T$, of size at most $k$. 
ALG is $\al$-approximately composable coreset if for any $T_i$,
\[
f(\bigcup(ALG(T_i)))\ge \al f_k(\bigcup T_i)
\]

?: $c$-composable coreset of size $k$: chunks of size $\sqrt{nk}$.

Mp-reduce model.

\begin{enumerate}
\item
Diversity maximization: maximize sum of pairwise distances diversity($S$).

Local search for diversity maximization (1 machine): Initialize $S$ with arbitrarily $k$ points which contains two fartest points. While there is a swap that improves diversity by $1+\fc{\ep}n$, then swap. This give 2-approximation.
%matroid constraints.

A local search algorithm computes a constant-factor composable coreset for diversity. Greedy algorithm computes  a 3-composable core-set.

%k-center 
\item
Distributed clustering

balanced version: size constraint on each cluster. 

Approx ratio: small constant $\cdot$ best single machine ratio.

Constant (2) rounds of MapReduce.

Experiments: (Capacitated $k$-center)

$10^7/10^8$ distances on US/world graph: Dividing into 300/1000 pieces, get 1.52/1.58 factor increase in OPT.
\item
Submodular function optimization: $f(A)+f(B)\ge f(A\cup B)+f(A\cap B)$, or $f(A\cup \{x\}) - f(A) \ge f(B\cup \{x\}) -f(B)$. (Diminishing return.)

Max-coverage problem is a special case. Applications: data summarization, data clustering, column selection, diversity maximization in search.

ML applications: exemplar based clustering, active set selections, graph cuts.

$f(S)=k$-median cost(empty set) - k-m

There exists no better than $\fc{\ln k}{\sqrt k}$ approximate composable core-set for submodular maximization. How about just max-cover?

None achieve constant factor in constant communication. Lose logarithmic in one of them. 

Randomization comes to rescue. $\al$-approximate randomized composable core-set iff
\[
\E[f_k(\bigcup ALG(T_i))] \ge \al \E[f(\bigcup T_i)].
\]
There exists a class of $O(1)$-approximate randomized composable coresets for monotone and non-monotone submodular maximization.
%
%Greedy algorithm: $\rc3$-approximate.

We show it for $\be$-nice algorithms: 
\begin{enumerate}
\item
$x\in T\bs ALG(T)$: ALG$(T)=$ALG($T\bs \{x\}$). ``Independence of irrelevant alternatives"
\item $\De(x,ALG(T))\le?$.
\end{enumerate}

Greedy is 1-nice.
%linear time clean IEEE.
\end{enumerate}•

$\rc3$ analysis.
\begin{enumerate}
\item
$OPT'=OPT\cap (\bigcup_i S_i)$, $OPT''=OPT\bs OPT'$.
\item
Linearize marginal contributions of elements in OPT: $\pi$ permutation on elements of OPT. $OPT^x=$ elements of OPT before $x$ in $\pi$. 
\item
Writing in terms of $\De$...
\item
%Lower bound $\max(f(OPT
\item
$(1-\rc{e})\rc3$.
%get better than 27\%?
\end{enumerate}•
Hard go beyond $\rc2$ with size $k$< impossible get beyond $1-\rc{e}$. .58-approximate randomized composable core-set of size $4k$.
If each machine outputs size $k'$, get $\sfc{k'}k$-approximate randomized composable coreset.

Graph mining frameworks
\begin{enumerate}
\item
iterative MapReduce
\item
Iterative MapReduce and DHT (distributed hash table) service (flume): fast random access to shared memory
\item
Pregel
\item
ASYMP (asynchronous message-passing). New way do some graph algorithms. Better than current optimized algorithms and using less resources.
\end{enumerate}