\def\filepath{C:/Users/Owner/Dropbox/Math/templates}
%\def\filepath{C:/Users/holden-lee/Dropbox/Math/templates}

\input{\filepath/packages_book.tex}
\input{\filepath/theorems_with_boxes.tex}
\input{\filepath/macros.tex}
\input{\filepath/formatting.tex}
\input{\filepath/other.tex}
\input{\filepath/theorem_num.tex}

\def\name{Functional Analysis}

\input{\filepath/titlepage.tex}

\pagestyle{fancy}
%\addtolength{\headwidth}{\marginparsep} %these change header-rule width
%\addtolength{\headwidth}{\marginparwidth}
\lhead{Part III}
\chead{} 
\rhead{Functional Analysis} 
\lfoot{} 
\cfoot{\thepage} 
\rfoot{} % !! Remember to change the problem set number
\renewcommand{\headrulewidth}{.3pt} 
%\renewcommand{\footrulewidth}{.3pt}
\setlength\voffset{0in}
\setlength\textheight{648pt}

\begin{document}
\title{Functional Analysis}
\author{Part III Course}
\date{Michaelmas 2013}
\maketitle

\tableofcontents

\chapter*{Introduction}
Andr\'as Zs\'ak (a.zsak@dpmms.cam.ac.uk) taught a course on Functional Analysis at
the University of Cambridge in Michaelmas (Fall) 2013. The website is \url{http://www.dpmms.cam.ac.uk/\~az10000}.  
These are my ``live-\TeX ed'' notes from the course.  The template was created by Akhil Mathew.

%\fixme{Questions will be in red.}

%\cpbox{Things to remember will be in boxes.}

%\wrbox{Things to watch out for will also be in boxes.}

We'll cover the main theorems that are useful in many other areas.

Here's an outline of the course.
\begin{enumerate}
\item
Introduction
\item
Hahn-Banach Theorem
\item
Riesz Representation Theorem (dual of continuous functions in a compact Hausdorff space)
\item
Weak topologies (weak and weak*, Banach--Alaoglu, etc.)
\item
Convexity and Krein-Milman Theorem
\item
Banach algebras
\item
Holomorphic functional calculus
\item
$C^*$-algebras (a type of Banach algebra)
\item
Borel functional calculus and spectral theory
\end{enumerate}
Here are some recommended books.
\begin{enumerate}
\item
G. R. Allan: Introduction to Banach Spaces and Algebras, OUP
\item
B. Bollobas: Linear Analysis.
\item
G. J. Murphy: $C^*$-algebras and operator theory
\item
W. Rudin: Real and Complex Analysis, Functional Analysis
\end{enumerate}
Prerequisites are linear analysis, complex analysis (section 7), and measure theory (sections 3, 9).


\input{functional_chapters/1.tex}
\input{functional_chapters/2.tex}
\input{functional_chapters/3.tex}
\input{functional_chapters/4.tex}
\input{functional_chapters/5.tex}
\input{functional_chapters/6.tex}
\input{functional_chapters/7.tex}
\input{functional_chapters/8.tex}
\input{functional_chapters/9.tex}
\appendix

\input{functional_chapters/a.tex}
\input{functional_chapters/b.tex}
\input{functional_chapters/c.tex}
\input{functional_chapters/summary.tex}

\bibliographystyle{plain}
\bibliography{\filepath/refs}

\printnomenclature
\printindex
\end{document}